---
title: "Spatial Correlation"
subtitle: "Applied Economics Research Course"
author: "Bas Machielsen"
aspectratio: 169
format: beamer
execute:
  freeze: auto
header-includes:
  - \usepackage{underscore}
  - \usepackage{booktabs}
---

```{r setup}
#| include: false
library(tidyverse)
library(sf)
library(spdep)
```

# Motivation: The Persistence Puzzle

## Historical Persistence in Economics

- A large literature asks: do historical events shape modern economic outcomes?
- Examples:
  - Colonial institutions → modern income (Acemoglu, Johnson & Robinson, 2001)
  - Slave trade → modern distrust (Nunn & Wantchekon, 2011)
  - Plough use → gender norms today (Alesina, Giuliano & Nunn, 2013)
  - Roman roads → modern economic activity
- These studies typically find **strong, statistically significant** results
- $t$-statistics of 4, 5, even 6 are common

## The Paradox

- Conley & Kelly (2025) notice something striking:

> "The historical persistence literature reports an unusual prevalence of large $t$-statistics."

- If history truly determines modern outcomes, some persistence is plausible...
- ...but $t$-statistics this large imply implausibly precise estimates
- Across **30 seminal papers**, rejection rates are far above what we'd expect
- **The claim:** standard errors may be systematically too small

## Why Might Standard Errors Be Too Small?

- Regression with geographic units: countries, regions, municipalities
- These units are **not independent**: nearby places share
  - Climate, geography, culture
  - Historical shocks
  - Institutional spillovers
- When errors are **spatially correlated**, observations carry less independent information than their count implies
- Classical OLS standard errors treat all observations as independent → understate uncertainty
- This is the **core problem** we will study today

# What is Spatial Correlation?

## Tobler's First Law of Geography

> "Everything is related to everything else, but near things are more related than distant things."

\vspace{0.5em}

- Waldo Tobler (1970) — arguably the most cited sentence in geography
- It is an empirical regularity, not a theorem
- But it holds remarkably widely:
  - Income, institutions, culture cluster geographically
  - Soils, climate, vegetation cluster geographically
  - Historical shocks (wars, plagues, colonisation) have spatial footprints

\vspace{0.5em}

**Implication for econometrics:** residuals from spatial regressions will typically be spatially correlated.

## Formal Definition

Consider a regression with $n$ geographic units:

$$
y_i = \mathbf{x}_i'\boldsymbol{\beta} + u_i, \quad i = 1, \ldots, n
$$

**Spatial correlation** means:

$$
\text{Cov}(u_i, u_j) \neq 0 \quad \text{for units } i, j \text{ that are spatially close}
$$

This is analogous to **serial correlation** in time series — but across space rather than time.

**Key difference from time series:** space is two-dimensional and has no natural ordering. There is no single "lag".

## The Spatial Weight Matrix

- To formalise "who is close to whom", we use a **spatial weight matrix** $W$
- $W$ is an $n \times n$ matrix where $w_{ij}$ encodes the spatial relationship between units $i$ and $j$
- Common choices:
  - **Contiguity:** $w_{ij} = 1$ if $i$ and $j$ share a border, 0 otherwise
  - **Distance-based:** $w_{ij} = d_{ij}^{-1}$ (inverse distance)
  - **$k$-nearest neighbours:** $w_{ij} = 1$ if $j$ is among $i$'s $k$ nearest neighbours

- By convention: $w_{ii} = 0$ (no self-neighbours)
- Usually **row-standardised**: each row sums to 1, so $\sum_j w_{ij} = 1$

## Spatial Lag

- The **spatial lag** of a variable $y$ is $Wy$
- Row-standardised: $(Wy)_i = \sum_j w_{ij} y_j$ — a weighted average of neighbours' values
- Spatial correlation means $y_i$ and $(Wy)_i$ are correlated
- This is directly analogous to $y_t$ and $y_{t-1}$ in time series

\vspace{0.5em}

**Intuition:** if your neighbours have high income, you probably do too. If the regression residual is positive for Paris, it is probably positive for Lyon too.

## Visualising Spatial Correlation

```{r spatial_noise_sim}
#| output-location: slide
#| fig-height: 4
#| fig-width: 9
#| fig-align: 'center'
#| cache: true
set.seed(42)
n <- 20
grid <- expand.grid(x = 1:n, y = 1:n)

# Pure random noise
grid$random <- rnorm(n^2)

# Spatially correlated: smooth + noise
grid$spatial <- as.vector(outer(sin(1:n * pi / n), sin(1:n * pi / n))) +
  rnorm(n^2, sd = 0.3)

p1 <- ggplot(grid, aes(x, y, fill = random)) +
  geom_tile() + scale_fill_viridis_c() +
  labs(title = "Pure random noise", fill = NULL) + theme_minimal()

p2 <- ggplot(grid, aes(x, y, fill = spatial)) +
  geom_tile() + scale_fill_viridis_c() +
  labs(title = "Spatially correlated", fill = NULL) + theme_minimal()

library(patchwork)
p1 + p2
```

## Why Does Geography Cluster?

Spatial correlation in regression residuals can arise from:

1. **Omitted variables** that are themselves spatially distributed (climate, soil quality, disease environment)
2. **Spatial spillovers**: outcomes in one location affect neighbouring locations (trade, migration, institutional imitation)
3. **Common shocks**: historical events affect entire regions (wars, epidemics, colonial borders)
4. **Measurement error** that is geographically structured

\vspace{0.5em}

**Important:** even after controlling for observed spatial covariates, residuals are often spatially correlated. This is the relevant concern for inference.

# Consequences for OLS

## OLS Assumptions Revisited

The standard OLS variance formula assumes **spherical errors**:

$$
E[\mathbf{u}\mathbf{u}'] = \sigma^2 \mathbf{I}_n
$$

Under this assumption:

$$
\widehat{\text{Var}}(\hat{\boldsymbol{\beta}}) = s^2 (\mathbf{X}'\mathbf{X})^{-1}
$$

When this holds, OLS is **BLUE** (Gauss-Markov theorem) and standard errors are valid.

## What Breaks with Spatial Correlation?

When $E[\mathbf{u}\mathbf{u}'] = \boldsymbol{\Sigma} \neq \sigma^2 \mathbf{I}$, the **true** variance of $\hat{\boldsymbol{\beta}}$ is:

$$
\text{Var}(\hat{\boldsymbol{\beta}}) = (\mathbf{X}'\mathbf{X})^{-1} \mathbf{X}'\boldsymbol{\Sigma}\mathbf{X} (\mathbf{X}'\mathbf{X})^{-1}
$$

But we compute $s^2(\mathbf{X}'\mathbf{X})^{-1}$ instead. The **ratio** depends on the off-diagonal elements of $\boldsymbol{\Sigma}$.

- If $\text{Cov}(u_i, u_j) > 0$ for nearby units (the typical case), classical SEs **understate** true uncertainty
- $t$-statistics are **inflated** — we reject $H_0$ too often
- OLS is still **unbiased** for $\boldsymbol{\beta}$, but **inference is invalid**

## Effective Sample Size

**Intuition:** imagine $n = 100$ municipalities, but due to strong spatial clustering they behave like 10 independent "blocs". Then:

- We effectively have $n_{\text{eff}} = 10$ independent observations, not 100
- Standard errors should scale as $1/\sqrt{10}$, not $1/\sqrt{100}$
- Classical SEs are too small by a factor of $\sqrt{10}$, inflating $t$-statistics by $\sqrt{10} \approx 3.2$

**General formula** for effective sample size under AR(1)-like spatial dependence with correlation $\rho$:

$$
n_{\text{eff}} \approx n \cdot \frac{1 - \rho}{1 + \rho}
$$

At $\rho = 0.5$: $n_{\text{eff}} \approx n/3$. At $\rho = 0.8$: $n_{\text{eff}} \approx n/9$.

## The Conley & Kelly (2025) Simulation

They demonstrate the problem starkly:

1. Draw a **"historical variable"** $x$: spatial noise with a north-south trend on a unit square ($n = 250$)
2. Draw an **independent "modern outcome"** $y$: independent spatial noise
3. Regress $y$ on $x$ — by construction, $\beta = 0$
4. Repeat many times

**Results without spatial correction:**

- Nominal 5% test rejects the true null **27% of the time**
- In one draw: $t = -3.8$, nominal $p = 0.0001$
- Placebo $p$-value (the true one): $p = 0.10$

**Lesson:** spatially structured data can generate convincingly significant results from pure noise.

## Rejection Rates Under Spatial Correlation {shrink=20}

| Method | $\rho = 0$ | $\rho = 0.3$ | $\rho = 0.5$ |
|---|---|---|---|
| HC (robust, no spatial correction) | 5% | 25% | 45% |
| Conley SEs (distance cutoff) | 5% | 10% | 15% |
| Large clusters (BCH, $k=6$) | 5% | 7% | 10% |
| Spatial basis regression | 5% | 6% | 7% |

*Source: Conley & Kelly (2025), Figure 2. Nominal size: 5%.*

All methods overreject, but **spatial basis regression** and **large cluster inference** perform best.

# Testing for Spatial Correlation: Moran's I

## Moran's I — Definition

Moran's $I$ is the most widely used test for **global** spatial autocorrelation:

$$
I = \frac{n}{S_0} \cdot \frac{\sum_i \sum_j w_{ij}(y_i - \bar{y})(y_j - \bar{y})}{\sum_i (y_i - \bar{y})^2}
$$

where $S_0 = \sum_i \sum_j w_{ij}$ is the sum of all weights.

- Proposed by Patrick Moran (1950)
- Defined for any variable $y$ and weight matrix $W$
- In practice, we apply it to **OLS residuals** to test for model misspecification

## Moran's I — Application to Residuals

Instead of the raw outcome variable, you plug in $\hat{u}_i$ — the OLS residual for unit $i$. Since OLS residuals sum to zero (when there's an intercept),  $\bar{y} = \bar{\hat{u}} = 0$, so the formula simplifies to:

$$
I = \frac{n}{S_0} \cdot \frac{\sum_i \sum_j w_{ij},\hat{u}_i,\hat{u}_j}{\sum_i \hat{u}_i^2}
$$                          


The weights $w_{ij}$ represent the same spatial weight matrix defined earlier in the slides (contiguity, distance-based, $k$-NN). 

## Moran's I — Intuition

Moran's $I$ is the spatial analogue of Pearson's correlation coefficient.

**Rewrite it as:**

$$
I \propto \frac{\sum_i \sum_j w_{ij}(y_i - \bar{y})(y_j - \bar{y})}{\sum_i (y_i - \bar{y})^2}
$$

This is a weighted sum of cross-products $(y_i - \bar{y})(y_j - \bar{y})$ — exactly what we compute in a correlation, but between unit $i$ and its spatial neighbours.

- $I > 0$: **positive spatial autocorrelation** — similar values cluster together
- $I < 0$: **negative spatial autocorrelation** — dissimilar values are neighbours (rare)
- $I \approx 0$: **spatial randomness**

## Moran's I — Inference

**Expected value under the null (spatial randomness):**

$$
E[I] = \frac{-1}{n - 1} \approx 0 \text{ for large } n
$$

**Variance under normality assumption:**

$$
\text{Var}(I) = \frac{n^2 S_1 - n S_2 + 3 S_0^2}{S_0^2(n^2 - 1)} - E[I]^2
$$

where $S_1 = \frac{1}{2}\sum_i \sum_j (w_{ij} + w_{ji})^2$, $S_2 = \sum_i \left(\sum_j w_{ij} + \sum_j w_{ji}\right)^2$.

**In practice:** use a **permutation test** — randomly permute labels across locations, recompute $I$ many times, compare observed $I$ to the permutation distribution.

## The Moran Scatterplot

- Plot $y_i - \bar{y}$ (horizontal axis) against the spatial lag $(Wy)_i - \bar{y}$ (vertical axis)
- The **slope** of the best-fit line is Moran's $I$
- Four quadrants:
  - **High-High (HH):** high value surrounded by high neighbours — spatial cluster
  - **Low-Low (LL):** low value surrounded by low neighbours — spatial cluster
  - **High-Low (HL):** high value surrounded by low neighbours — spatial outlier
  - **Low-High (LH):** low value surrounded by high neighbours — spatial outlier

The scatterplot decomposes the global $I$ into local contributions.

## Computing Moran's I in R — Data

```{r netherlands_data}
#| include: false
nl <- st_read("../../data/netherlands/netherlands_roman_updated.geojson", quiet = TRUE)
```

```{r moran_weights}
#| echo: true
#| eval: true
#| cache: true
# Build spatial weight matrix from contiguous neighbours
nb <- poly2nb(nl, queen = TRUE)                    # queen contiguity
W  <- nb2listw(nb, style = "W", zero.policy = TRUE) # row-standardise
```

## Computing Moran's I in R — Test {shrink=20}

```{r moran_test}
#| echo: true
#| eval: true
#| output-location: slide
# Test Moran's I on distance to Roman Empire border
moran.test(nl$distance_to_border, W, zero.policy = TRUE)
```

## Moran Scatterplot in R

```{r moran_plot}
#| echo: true
#| eval: true
#| output-location: slide
#| fig-height: 4
#| fig-width: 6
#| fig-align: 'center'
moran.plot(
  nl$distance_to_border,
  W,
  xlab = "Distance to Roman border",
  ylab = "Spatial lag (neighbours' average)",
  main = "Moran Scatterplot — Netherlands municipalities",
  zero.policy = TRUE
)
```

## Implications of Moran's I

**For variable selection:**

- High $I$ in $y$ or $x$ alone does not invalidate OLS — we care about $I$ in **residuals**

**For residual diagnostics:**

- After fitting a model, compute Moran's $I$ on residuals $\hat{u}_i$
- Significant $I$ → spatial patterns remain unexplained → SEs unreliable

**For interpretation:**

- Moran's $I$ is a **global** summary — it detects overall clustering
- It does **not** tell you where the clusters are (for that: LISA maps)
- A non-significant $I$ does not rule out local spatial patterns

## Moran's I on Regression Residuals

```{r moran_residuals}
#| echo: true
#| eval: true
#| output-location: slide
#| cache: true
# Simple OLS: does Roman Empire proximity predict municipal area?
model <- lm(area ~ distance_to_border, data = nl)
residuals_nl <- residuals(model)

# Moran's I on residuals
moran.test(residuals_nl, W, zero.policy = TRUE)
```

# Other Spatial Diagnostics

## Geary's C

An alternative to Moran's $I$, focusing on **local** differences:

$$
C = \frac{(n-1)\sum_i \sum_j w_{ij}(y_i - y_j)^2}{2 S_0 \sum_i (y_i - \bar{y})^2}
$$

- $C \in (0, 2)$; expected value under null: $C = 1$
- $C < 1$: positive spatial autocorrelation
- $C > 1$: negative spatial autocorrelation
- More sensitive to **local** spatial autocorrelation than Moran's $I$
- Moran's $I$ and Geary's $C$ are approximately inversely related: $I \approx 1 - C$

## Local Moran's I (LISA)

**Local Indicators of Spatial Association** (Anselin, 1995) decompose global $I$ into unit-level contributions:

$$
I_i = \frac{(y_i - \bar{y})}{m_2} \sum_j w_{ij}(y_j - \bar{y}), \quad m_2 = \frac{1}{n}\sum_i (y_i - \bar{y})^2
$$

Note: $I = \sum_i I_i / S_0$.

**Uses of LISA:**
- Identify specific **spatial clusters** (HH, LL) and **outliers** (HL, LH)
- Detect local non-stationarity — global $I \approx 0$ but local clusters exist
- Map significance of each unit's $I_i$ after multiple-testing correction

# Classical Correction: Conley Standard Errors

## The Idea

- Allow errors to be correlated across units within a **distance bandwidth** $d$
- Beyond $d$: errors treated as uncorrelated
- A **HAC** (heteroskedasticity and autocorrelation consistent) estimator for spatial data

The sandwich variance estimator:

$$
\widehat{\text{Var}}(\hat{\boldsymbol{\beta}}) = (\mathbf{X}'\mathbf{X})^{-1} \hat{\mathbf{S}} (\mathbf{X}'\mathbf{X})^{-1}
$$

$$
\hat{\mathbf{S}} = \sum_i \sum_j K\!\left(\frac{d_{ij}}{d}\right) \hat{u}_i \hat{u}_j \mathbf{x}_i \mathbf{x}_j'
$$

where $K(\cdot)$ is a kernel (e.g., Bartlett) that downweights distant pairs.

- Proposed by Conley (1999) — a standard tool in spatial econometrics

## Limitations of Conley Standard Errors

Conley SEs are **not sufficient** (Conley & Kelly, 2025):

- Require choosing a bandwidth $d$ — results sensitive to this choice
- Perform poorly when spatial correlation is **large** or **long-range**
- Assume the spatial correlation structure decays with distance — may not hold for trends
- At $\rho = 0.5$: rejection rate is **15%** vs nominal 5%

**Root cause:** Conley SEs correct the variance estimator but do not address **spatial trends** in the data. If $x$ and $y$ both have a north-south gradient, their correlation looks significant even if the gradient in $y$ has nothing to do with $x$.

# Conley & Kelly (2025): Spatial Basis Regression

## The Core Idea

**Problem:** $x$ and $y$ may share common spatial trends that create spurious correlation.

**Solution:** control for those trends directly by adding **flexible spatial controls** to the regression.

$$
y_i = \beta x_i + f(\text{lon}_i, \text{lat}_i) + \varepsilon_i
$$

where $f$ is a smooth function of coordinates. The function absorbs:
- North-south gradients
- East-west gradients
- Other spatial trends

After removing spatial trends from both $y$ and $x$, the residual variation used to identify $\beta$ is less spatially correlated.

## Tensor Product B-Splines

Conley & Kelly use **tensor product B-splines** to estimate $f$:

$$
f(\text{lon}, \text{lat}) = \sum_{j=1}^{J} \sum_{k=1}^{K} \tau_{jk} \, a_j(\text{lon}) \cdot b_k(\text{lat})
$$

- $a_j(\cdot)$ and $b_k(\cdot)$ are **B-spline basis functions** (piecewise polynomial)
- The number of basis functions $J \times K$ controls flexibility
- **Basis dimension selected by BIC** — avoids overfitting while controlling for relevant trends

**Interpretation:** we are partitioning out a smooth spatial surface from both the outcome and the treatment variable. What remains is the within-surface variation.

## Why Splines Work: An Analogy

Think of **regional fixed effects** — controlling for broad geographic areas:

- Absorbs common shocks within regions
- But regions are discrete; spatial correlation is continuous

B-splines generalise this:
- Absorb spatial trends **continuously** across the map
- More flexible than dummies, less arbitrary than polygons
- BIC ensures we don't over-absorb (which would consume the treatment's variation)

This is similar to adding **time trends** in a panel model to control for secular trends that affect all units.

## Implementation in R {shrink=20}

```{r basis_regression}
#| echo: true
#| eval: true
#| cache: true
# Extract coordinates
coords <- st_centroid(nl) |> st_coordinates()
nl$lon <- coords[, 1]
nl$lat <- coords[, 2]

# Spatial basis regression: add polynomial in (lon, lat)
# (In practice Conley & Kelly use B-splines via splines::bs())
model_ols   <- lm(area ~ distance_to_border, data = nl)
model_basis <- lm(area ~ distance_to_border +
                    poly(lon, lat, degree = 3),
                  data = nl)

# Compare Moran's I on residuals
cat("Moran's I — without basis:",
    round(moran.test(residuals(model_ols),   W, zero.policy = TRUE)$statistic, 3), "\n")
cat("Moran's I — with basis:   ",
    round(moran.test(residuals(model_basis), W, zero.policy = TRUE)$statistic, 3), "\n")
```

## What the Spatial Basis Achieves

After adding basis terms, Conley & Kelly (2025) find:

- **Moran statistics drop substantially** in residuals of most persistence studies
- t-statistics fall from, e.g., 5.25 to 1.5 (Becker & Woessmann study)
- The dramatic significance of many findings largely disappears

**Key insight:** the historical variable and the modern outcome often share the **same spatial surface** — a confound, not a causal mechanism. The spatial basis absorbs this confound.

# Large Cluster Inference

## BCH: Bester, Conley & Hansen (2011)

**Idea:** divide the $n$ observations into $k$ large **spatial clusters**, then apply cluster-robust inference.

- Use **$k$-medoids clustering** on coordinates to form geographic clusters
- Apply the cluster-robust variance estimator (Liang & Zeger, 1986) treating each cluster as a unit
- Each cluster acts like a "super-observation" — within-cluster correlation is unrestricted

**Requirements:**
- $k \geq 6$ clusters for reliable inference (need enough "units" for asymptotics)
- Clusters should be large enough to capture spatial correlation within them
- Tradeoff: more clusters → less conservative but more assumptions

## IM: Ibragimov & Müller (2010)

**A different approach:** estimate $\hat\beta$ separately within each cluster, then test $H_0: \beta = 0$ using the **distribution of cluster-specific estimates**.

1. Divide observations into $k$ spatial clusters
2. Estimate $\hat\beta_1, \ldots, \hat\beta_k$ within each cluster
3. Test $H_0: \mu = 0$ where $\mu = E[\hat\beta_c]$ using a two-sided $t$-test with $k-1$ degrees of freedom

**Advantage:** valid under very general within-cluster dependence. No assumptions about the correlation structure.

**Advantage:** intuitive — if the treatment effect is real, it should appear in most clusters. If only some clusters drive the result, that is informative.

## Comparing the Approaches {shrink=20}

| Method | Key idea | Pros | Cons |
|---|---|---|---|
| HC SEs | Robust to heteroskedasticity | Simple | Ignores spatial correlation |
| Conley SEs | HAC with distance kernel | Well-known | Bandwidth choice; insufficient for trends |
| BCH | Large geographic clusters | No correlation assumptions | Needs $k \geq 6$; BIC choice |
| IM | Cluster-level $t$-test | Intuitive; very robust | Low power with few clusters |
| Spatial basis | Remove spatial trends + HC | Addresses root cause | May reduce power; BIC choice |

**Recommendation:** use spatial basis regression as the default; corroborate with BCH or IM.

# Diagnostic Tests

## The Placebo Treatment Test

**Core idea:** if the finding is real, shuffling the treatment's spatial locations should destroy the result.

**Procedure:**
1. Estimate the model, record $\hat\beta_{\text{obs}}$
2. Randomly permute the coordinates (or the treatment variable) across units
3. Re-estimate the model on the permuted data, record $\hat\beta_{\text{perm}}$
4. Repeat many times (e.g., 1000) to build the **empirical null distribution**
5. The **placebo $p$-value** is the share of permutations where $|\hat\beta_{\text{perm}}| \geq |\hat\beta_{\text{obs}}|$

**Why it works:** permutation destroys the spatial structure of the treatment while preserving the spatial structure of the outcome. If the coefficient survives permutation, the spatial trend in $x$ is correlated with the spatial trend in $y$ even when we break the unit-level link.

## The Synthetic Outcome Test

**Core idea:** generate fake outcomes that mimic the spatial structure of the true outcome, then check whether the observed $\hat\beta$ is unusual.

**Procedure:**
1. Estimate a **spatial noise model** for $y$: $\hat{y}_i = \hat{f}(\text{lon}_i, \text{lat}_i) + \hat\varepsilon_i$
2. Simulate many synthetic outcomes $y^*$ from this model (matching the spatial trend and noise structure)
3. Regress each $y^*$ on the actual treatment $x$ → synthetic coefficient distribution
4. Check where $\hat\beta_{\text{obs}}$ falls in this distribution

**Advantage over placebo test:** tests against the appropriate null (data-generating process), not just spatial randomisation.

## Conley & Kelly Apply This to 30 Papers {shrink=20}

They apply spatial basis regression, BCH, IM, and placebo tests to **30 seminal historical persistence papers**:

| Correction method | Studies losing significance at 5% |
|---|---|
| None (original) | 0 / 30 |
| Conley SEs (generous bandwidth) | ~5 / 30 |
| Spatial basis + HC | ~15 / 30 |
| Placebo test | ~15 / 30 |
| BCH ($k = 4$) | ~12 / 30 |
| IM ($k = 4$) | ~10 / 30 |

*Approximate figures from Table 1 and Figures 3–4 of Conley & Kelly (2025).*

**Takeaway:** roughly half of the literature's findings become insignificant under appropriate correction.

## Interpreting the Placebo Results

- A high placebo $p$-value does **not** prove the effect is spurious
- It means: the spatial structure of $x$ is consistent with producing the observed $\hat\beta$ by chance
- A low placebo $p$-value provides stronger evidence for a genuine effect

**Why the persistence literature is particularly vulnerable:**

- Historical variables (Roman roads, colonial institutions, settler mortality) are **strongly spatially structured**
- Modern outcomes (income, trust, gender norms) are also **strongly spatially structured**
- Both share climate, geography, and historical trajectories — a rich confound structure

# Practical Guidance

## A Checklist for Spatial Regressions

\vspace{0.5em}

**Step 1 — Visualise:**
- Map your outcome variable and key regressors
- Map OLS residuals — look for spatial patterns

**Step 2 — Test:**
- Compute Moran's $I$ on residuals with a reasonable weight matrix
- If $I$ is significant, spatial correlation is present

**Step 3 — Correct:**
- Add spatial basis terms (polynomial or B-splines in lon/lat) selected by BIC
- Report results with and without spatial basis

**Step 4 — Validate:**
- Run BCH or IM as a robustness check
- Run placebo treatment test
- If results are robust to all corrections, you have stronger evidence

## What "Correcting" Does and Does Not Do

**Does:**
- Provide valid inference in the presence of spatial correlation
- Reduce inflated $t$-statistics to appropriate levels
- Distinguish genuine effects from spurious spatial correlations

**Does not:**
- Solve endogeneity (use IV for that)
- Guarantee a causal interpretation
- Replace the need for a credible identification strategy

**Spatial basis regression absorbs the confound of shared spatial trends.** It is not a substitute for — but a complement to — good research design.

## When to Worry Most

Spatial correlation is a bigger concern when:

- The treatment variable has strong **geographic clustering** (colonial-era variables, climate zones, borders)
- The sample covers a **large geographic area** with potential for long-range trends
- There are few natural **discontinuities** to exploit (RDD-based designs are more robust)
- The outcome is a modern aggregation of processes that were themselves spatially structured

**Reassuring cases:** geographic RDD designs with tight bandwidths near a border naturally control for spatial trends — the basis regression is less critical.

# Conclusion

## Summary

**The problem:**
- Spatially structured data violates the i.i.d. assumption
- OLS standard errors are too small → inflated $t$-statistics
- Effect is quantitatively large: nominal 5% tests reject at 27–45% in simulations

**The diagnostic:**
- Moran's $I$ on residuals detects spatial correlation
- Moran scatterplot reveals the nature of spatial dependence

**The solutions (Conley & Kelly, 2025):**
1. **Spatial basis regression** — remove shared spatial trends with flexible controls
2. **Large cluster inference** (BCH, IM) — treat geography as the clustering variable
3. **Placebo and synthetic outcome tests** — validate the spatial robustness of findings

## Key References

- Conley, T.G. (1999). GMM Estimation with Cross Sectional Dependence. *Journal of Econometrics*.
- Tobler, W. (1970). A Computer Movie Simulating Urban Growth in the Detroit Region. *Economic Geography*.
- Moran, P.A.P. (1950). Notes on Continuous Stochastic Phenomena. *Biometrika*.
- Anselin, L. (1995). Local Indicators of Spatial Association -- LISA. *Geographical Analysis*.
- Bester, C.A., Conley, T.G. & Hansen, C.B. (2011). Inference with Dependent Data Using Cluster Covariance Estimators. *Journal of Econometrics*.
- Ibragimov, R. & Müller, U.K. (2010). $t$-statistic Based Correlation and Heterogeneity Robust Inference. *Journal of Business & Economic Statistics*.
- **Conley, T.G. & Kelly, M. (2025). Standard Errors of Persistence. *Journal of International Economics*.**

# Application: Chapman (2026)

## Context: Democracy and Redistribution

**Chapman, J. (2026). "Democracy, Redistribution, and Inequality: Evidence from the English Poor Law." *Journal of the European Economic Association*.**

\vspace{0.3em}

**The question:** Does income inequality shape how much democratisation increases redistribution?

- The **Meltzer-Richard (1981) model**: the median voter, poorer than the mean, demands more redistribution as inequality rises
- But testing this empirically is hard: inequality, democracy, and redistribution are all endogenous

**The setting:** 19th-century England. Local poor law unions (boards of guardians) controlled spending on poor relief -- the main safety net before the modern welfare state. Funded by a proportional tax on rental values: the mechanism in Meltzer-Richard holds almost exactly.

## The Identification Strategy

**The exogenous shock:** The 1894 Local Government Act (LGA) democratised all poor law union boards simultaneously, removing four elite protections in one stroke:

1. Property qualifications to stand as guardian
2. A graduated franchise (wealthy voters held up to 12 votes)
3. Unelected *ex officio* magistrates (local landowners) on boards
4. The absence of a secret ballot

**Interpretation:** Because all boards were treated at the same time and the reform was driven by national political logic -- not local spending trends -- the LGA is as good as exogenous to each union.

**Empirical specification** (difference-in-differences with heterogeneous treatment):

$$
y_{it} = \beta \cdot \text{Inequality}_i \times \text{post1894}_t + \gamma_0 X'_{it} + T_t + \alpha_i + \varepsilon_{it}
$$

$\beta > 0$ means more unequal unions increased spending more after democratisation.

## Inequality Measure and Key Variables

**Measuring the Meltzer-Richard mechanism directly:**

$$
\text{meanMedianRatio}_i = \frac{\text{Mean rental value}}{\text{Median wage}}
$$

- Directly captures the relative tax burden of rich vs. poor (the Meltzer-Richard channel)
- Tax was levied on rental values, so this ratio maps precisely to the model
- Measured pre-reform to avoid endogeneity

**Other inequality dimensions tested (Table 3):**

| Channel | Measure | Predicted sign |
|---|---|---|
| Meltzer-Richard | Mean/median ratio | $+$ |
| Elite de facto power | Ex officio chairman | $-$ |
| Fairness concerns | Land concentration, large farms | $+$ |
| Economic need | Cereal suitability, \% agricultural laborers | $\pm$ |

## Addressing Spatial Correlation {shrink=20}

Income inequality is **geographically clustered** in this data -- arable agriculture (high inequality) dominates the east, pastoral (low inequality) the west. This is a textbook case for spatial concern.

**Chapman's approach, following Conley & Kelly (2025):**

1. **Moran's I diagnostic:** Computed for the inequality measure and for changes in relief spending. After controlling for agriculture type and other observables, *little evidence of spatial autocorrelation remains*.

2. **Conley standard errors:** All regressions use SEs that allow for arbitrary spatial correlation within a **100 km bandwidth** (Colella et al., 2019). Results are robust across a range of bandwidths.

3. **Flexible spatial time trends:** Specification (7) in Table 2 includes interactions of quartic time trends with longitude and latitude -- equivalent to a spatial basis regression -- and results are unchanged.

**Lesson:** spatial correlation was a genuine concern here, and the paper deals with it transparently and rigorously.

## Main Results

**Support for the Meltzer-Richard hypothesis (Table 2):**

- $\hat\beta \approx 0.11$ (SD units) across all specifications, significant at 1\%
- At the 90th percentile of inequality: spending rose by $\sim$13\% relative to 1893 levels
- Effect is concentrated on **outdoor relief** (direct cash/kind transfers), not indoor (workhouse) -- consistent with the redistributive mechanism
- Robust to crop controls, demographic controls, financial controls, union-specific time trends, and spatial corrections

**Elite de facto power (Table 3):**

- Unions with an unelected (ex officio) chairman before reform saw **smaller** spending increases ($\hat\beta \approx -0.13$)
- Elites used local institutional control to blunt democratisation's redistributive impact
- Effect mainly on less publicly salient spending items -- harder for voters to monitor

**Conclusion:** the paper provides clean empirical support for Meltzer-Richard in a historically controlled setting, while explicitly addressing spatial correlation as required by Conley \& Kelly.
