---
title: "Historical Persistence"
subtitle: "Applied Economics Research Course"
author: "Bas Machielsen"
institute: "Utrecht University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=F}
library(tidyverse); library(pacman); library(sf)
knitr::opts_chunk$set(warning=F, message=F, fig.retina = 3)
```

class: inverse, center, middle

# Introduction

---

# Introduction

- In the previous class, I have shown you an example of _vector_-format spatial data

- This means that spatial data is represented in a `data.frame`-like format with objects like polygons or lines
  - Polygons: Municipality boundaries
  - Lines: Borders of the Roman Empire
  
- Another way to represent spatial data is _raster_ data

- Raster data can be seen much like an image, consisting of a matrix of "pixels"
  - But these pixels have specific geographic coordinates indicating their place on the world globe
  
---

# Outline

- In this lecture, I demonstrate:

  - Part 1: Manipulate and join vector data
  
  - Part 2: Manipulate raster data
  
  - Part 3: How to use _vector data_ when you have rastered data as basic unit of analysis
  
  - Part 4: How to use _raster data_ when you have vector data as basic unit of analysis

---

# Why different formats?

- In a research project, you must pick your **unit of analysis**

- A unit of analysis in a research project with a spatial dimension can be either a geographical unit such as a _municipality_, a _region_ or a _country_
  - This naturally fits well with _vector_ data, containing polygons
  
- However, your unit of analysis may also be a _1x1 latitude x longitude_ area
  - Raster data is much more suitable for this
  
---

# Preliminaries

- Both raster and vector data have various things in common

  - Both are geocoded and use a particular _CRS_ (Coordinate Reference System)
  
  - Can have multiple variables (vector data) and layers (raster data)

- You can generally convert data from vector to raster

- Converting from raster to vector is more difficult

  - Usually what we do is use either raster data as a basis, or aggregate raster data to an already existing vector basis
  - Example: aggregate nightlights data (raster data) to the country level
    - See [This Paper](https://thedocs.worldbank.org/en/doc/350051528721174623-0050022018/original/Nightlights.pdf)
  
---


class: inverse, center, middle

# Part 1: Computing on Vector Data

---

# Getting the CRS

- You have already seen you can do various things on vector data

- For example, you can ask what _CRS_ it is in:

```{r download_netherl, cache=TRUE}
germany <- geodata::gadm("Netherlands", level=1, path="./") |> st_as_sf()
st_crs(germany)[1]
```

- (You can also change that)

- [R-Spatial](https://r-spatial.github.io/sf/articles/sf4.html) has a very good introduction to manipulating spatial data. In particular, it details how to:
  - Aggregate feature sets
  - Summarize feature sets
  - Join two feature sets based on feature geometry

---

# Spatially Merging Two Vector Data Sets


---

class: inverse, center, middle

# Part 2: Computing on Raster Data

---

# Computing on raster data


- You cannot do so much on raster data, however

- One thing you can do is _aggregate_ your data to a higher level of detail
  - (Or a lower level of granularity)
  
- I will demonstrate that here using a simple example: 


---

class: inverse, center, middle

# Part 3: Aggregate Raster to Vector

---

# Still

- Still do this part

- Use the nightlights data

---

class: inverse, center, middle

# Part 4: Aggregate Vector to Raster


---

# Download Map to Start With

- We first download the map of the Netherlands through the `giscoR` package
  
```{r get_municipalities, cache=TRUE}
netherlands <- giscoR::gisco_get_lau(year='2020', country='Netherlands')

netherlands |> head(5)
```


---

# Plot The Map

- Let us have a look at `POP_DENS_2020`, the population density in 2020
  - This is but one of the variables in this spatial data.frame
  
```{r plot_nl, fig.align='center', fig.height=6, fig.width=6}
netherlands |> 
  ggplot(aes(fill=log(POP_DENS_2020))) +
  geom_sf()
```

---

# Make Raster of The Map

- Next, we make a grid dividing the Netherlands into areas of $x \times y$ latitude x longitude
  - Say $0.05 \times 0.05$ in this case
  
- The grid literally consists of pieces of area of 0.05 latitude by 0.05 longitude
  - This can be the unit of our analysis

```{r make_grid}
library(sf); library(stars); library(starsExtra)

grid <- st_make_grid(netherlands, square=T, cellsize=c(0.05, 0.05))
```


---

# Aggregate Original Vector Data to Raster

- Let us now use `raster::aggregate` to `aggregate` the variables of `netherlands` to the grid-level

- We can choose an _aggregation function_: this time, we use the mean
  - That is, for each box in the grid, we compute the corresponding `mean` values of the polygons and compute a geographical "weighted average" of our population density and other variables

```{r aggregate_grid}
per_grid <- raster::aggregate(netherlands, grid, FUN=mean)
```

---

# Look at Plot

- Let us look at the output:

```{r plot_gridded, fig.height=6, fig.align='center'}
per_grid |> 
  st_as_sf() |> 
  ggplot(aes(fill=POP_DENS_2020)) + geom_sf()
```


---

class: inverse, center, middle

# Extra: `geodata`

---

# The `geodata` package

- The `geodata` package is a package through which you can get all kinds of open-sourced georeferenced data

- Install it and load it through `pacman`:

```{r geodataload}
p_load(geodata)
```

- For example, geographical data can be downloaded through:

```{r download_nl}
nl <- geodata::gadm("Netherlands", level = 2, path = "./") |> st_as_sf()
```

---

# The `geodata` package

- You can also download other kinds of (raster and vector) data:
  - This can again be aggregate to our raster or vector data!

```{r download_temp, fig.height=4, fig.align='center'}
temperature <- geodata::worldclim_country("Netherlands",var = "tavg", path = "./")
temperature$NLD_wc2.1_30s_tavg_1 |> plot()
```


```{r eval=F, include=F}
netherlands_raster <- st_rasterize(netherlands |> 
                                     dplyr::select(POP_DENS_2020)
                                   )

coords <- st_coordinates(netherlands_raster) |> 
  summarize(xmin = min(x), xmax = max(x),
            ymin = min(y), ymax = max(y))
```

