---
title: "Historical Persistence"
subtitle: "Applied Economics Research Course"
author: "Bas Machielsen"
institute: "Utrecht University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=F}
library(tidyverse); library(pacman); library(sf)
knitr::opts_chunk$set(warning=F, message=F, fig.retina = 3)
```

class: inverse, center, middle

# Introduction

---

# Introduction

- In the previous class, I have shown you an example of _vector_-format spatial data

- This means that spatial data is represented in a `data.frame`-like format with objects like polygons or lines
  - Polygons: Municipality boundaries
  - Lines: Borders of the Roman Empire
  
- Another way to represent spatial data is _raster_ data

- Raster data can be seen much like an image, consisting of a matrix of "pixels"
  - But these pixels have specific geographic coordinates indicating their place on the world globe
  
---

# Outline

- In this lecture, I demonstrate:

  - Part 1: Manipulate and join vector data
  
  - Part 2: Manipulate raster data
  
  - Part 3: How to use _vector data_ when you have rastered data as basic unit of analysis
  
  - Part 4: How to use _raster data_ when you have vector data as basic unit of analysis

---

# Why different formats?

- In a research project, you must pick your **unit of analysis**

- A unit of analysis in a research project with a spatial dimension can be either a geographical unit such as a _municipality_, a _region_ or a _country_
  - This naturally fits well with _vector_ data, containing polygons
  
- However, your unit of analysis may also be a _1x1 latitude x longitude_ area
  - Raster data is much more suitable for this
  
---

# Preliminaries

- Both raster and vector data have various things in common

  - Both are geocoded and use a particular _CRS_ (Coordinate Reference System)
  
  - Can have multiple variables (vector data) and layers (raster data)

- You can generally convert data from vector to raster

- Converting from raster to vector is more difficult

  - Usually what we do is use either raster data as a basis, or aggregate raster data to an already existing vector basis
  - Example: aggregate nightlights data (raster data) to the country level
    - See [this paper](https://thedocs.worldbank.org/en/doc/350051528721174623-0050022018/original/Nightlights.pdf)
  
---


class: inverse, center, middle

# Part 1: Computing on Vector Data

---

# Getting the CRS

- You have already seen you can do various things on vector data

- For example, you can ask what _CRS_ it is in:

```{r download_netherl, cache=TRUE}
netherlands <- geodata::gadm("Netherlands", level=1, path="./") |> st_as_sf()
st_crs(netherlands)[1]
```

- (You can also change that with `st_transform`)

- [R-Spatial](https://r-spatial.github.io/sf/articles/sf4.html) has a very good introduction to manipulating spatial data. In particular, it details how to:
  - Aggregate feature sets
  - Summarize feature sets
  - Join two feature sets based on feature geometry

---

# Spatial Data Operations

- `sf` allows you to overlay spatial objects to find their intersections, unions, or differences.
- Common operations include:
  - **Intersection**: Finding common areas between two spatial objects. (`st_intersection`)
  - **Union**: Combining the geometries of two or more objects. (`st_union`)
  - **Difference**: Identifying the areas where one object differs from another. (`st_difference`)
  
- You can also use *buffering*: buffering involves creating a buffer (a zone or area) around a spatial object.
- Useful for proximity analysis and determining distances to specific features.

```{r demo}
utrecht <- netherlands |> filter(NAME_1 == "Utrecht")
buffer_around_utrecht <- st_buffer(utrecht,
                                   dist = 1000)
```

---

# Clipping and Cropping

- Clipping is the process of extracting a subset of spatial data within a defined boundary.

```{r clip}
clipped_data <- st_intersection(netherlands, utrecht)
```

- Cropping is similar but keeps only the part of the data within a specific polygon.

```{r crop}
cropped_data <- st_crop(netherlands, utrecht)
```

```{r 2plots, fig.align='center', fig.width=8, fig.height=3}
library(gridExtra)
p1 <- ggplot(clipped_data) + geom_sf(); p2 <- ggplot(cropped_data) + geom_sf()
grid.arrange(p1, p2, ncol=2)
```

---

# Spatial Joins

- `sf` supports spatial joins to combine attribute data based on the spatial relationship.
- Types of spatial joins include:

- **Inner Join**: Keep only matching records from both datasets.
  
```{r injoin, eval=F}
inner_join_result <- st_join(data1, data2)
```
  
- **Left Join**: Include all records from the left dataset and matching records from the right.
  
```{r ljoin, eval=F}
left_join_result <- st_join(data1, data2, left = TRUE)
```

- **Right Join**: Include all records from the right dataset and matching records from the left.
  
```{r rjoin, eval=F}
right_join_result <- st_join(data1, data2, right = TRUE)
```

- **Full Join**: Include all records from both datasets.
  
```{r fj, eval=F}
full_join_result <- st_join(data1, data2, join = st_nearest_feature)
```


---

# Spatially Merging Two Vector Data Sets

- I will now demonstrate a very common approach to merging two vector data sets:

- Suppose you have one **province**-level dataset and one **municipality**-level dataset

- You want to merge them together, so that you retain the lowest level (municipality)
  - But you want your data to also contain provinces, so you can go back and forth
  
- Download the two maps using the `geodata` package
  - Convert them to `sf` format using `st_as_sf`

```{r download_munip}
municipalities <- geodata::gadm('Netherlands', level=2, path='./') |> 
  st_as_sf() |>
  select(NAME_2, geometry)

provinces <- geodata::gadm('Netherlands', level=1, path='./') |> 
  st_as_sf() |> 
  select(NAME_1, geometry)
```

---

# Inspecting the data

- The datasets look as follows (I only show municipality):

```{r show_munip}
municipalities |> head(5)
```

---

# Spatial joins

- Just like ordinary `join`'s, such as `left_join`, `full_join`, etc., you can also spatially join several datasets

- This is done using the `st_join` command:
  - You can specify various types of `join`
  - This time, we use `st_covered_by`: a municipality (the first object) should be 'covered by' a province (the second object)

```{r merge_munip}
merged_data <- st_join(municipalities, provinces, join=st_covered_by)
```

---

# Spatial join plot

- We can see that we have recovered our provinces (almost) perfectly


```{r plt, fig.width=5, fig.height=4, fig.align='center'}
merged_data |> 
  ggplot(aes(fill=NAME_1)) + geom_sf()
```

---

# Spatial join plot 

- ..The inaccuracy is easy to solve:

```{r mdcorr}
merged_data <- merged_data |> 
  mutate(NAME_1 = if_else(
    is.na(NAME_1), "Zuid-Holland", NAME_1
  ))
```

- The source of the inaccuracy was the missing name for Zuid-Holland in the `provinces` spatial data.frame

---

# Join spatial data with other data.frames

- You have already seen you can use `mutate` on a spatial data.frame, just as you would on a normal data.frame

- In fact, you can use this spatial data.frame as you would use any other data.frame

- For example, you can merge it with other data sets in exactly the way you're used to
  - Use the `cbsodataR` package:

```{r downloadpop}
library(cbsodataR)
population <- cbsodataR::cbs_get_data('85385NED') |>
  select(Naam_2, Inwonertal_54) |>
  mutate(Naam_2 = str_trim(Naam_2))
```

```{r dwpm}
data_with_pop <- left_join(merged_data, 
                           population,
                           by=c("NAME_2" = "Naam_2"))
```

---

# Plot the resulting map

- Now, we can plot the resulting map
  - Some provinces have been omitted because of faulty matches
  - Generally, you need _identifiers_ to match two data.frames without errors
  
```{r plotpop, fig.width=5, fig.height=4, fig.align='center'}
data_with_pop |> 
  ggplot(aes(fill=log(Inwonertal_54))) +
  geom_sf() +
  scale_fill_viridis_b()
```

---

class: inverse, center, middle

# Part 2: Computing on Raster Data

---

# Computing on raster data


- You cannot do so much on raster data, however

- One thing you can do is _aggregate_ your data to a higher level of detail
  - (Or a lower level of granularity)
  
- I will demonstrate that here using a simple example: 


---

class: inverse, center, middle

# Part 3: Aggregate Raster to Vector

---

# Raster Data

- We will use the MODIS *(Moderate Resolution Imaging Spectroradiometer)* satellite data source to acquire a raster data of the Netherlands

- MODIS is an instrument aboard the Terra and Aqua satellites, which orbits the entire Earth every 1-2 days, acquiring data at different spatial resolutions.

- The data acquired by MODIS describes features of the land, oceans and the atmosphere. 
  - A complete list of MODIS data products can be found on the [MODIS website](https://modis.gsfc.nasa.gov/data/dataprod/)
  - The website contains _codes_ that you use to query the corresponding data
  - I use the _vegetation index products_, available [here](https://modis.gsfc.nasa.gov/data/dataprod/mod13.php)

---

# Get vegatation data

- We start off by gathering the Normalized Difference Vegetation Index (NDVI) data for the Netherlands
  - It is a widely used vegetation index in remote sensing and geospatial analysis to assess and monitor the health and vitality of vegetation
  - Values close to -1 represent non-vegetated or barren surfaces, such as water bodies or urban areas, values close to 0 correspond to to bare soil, rocks, or other non-vegetated surface, and values closer to 1 represent various stages of vegetation, with higher values indicating healthier and denser vegetation.
    
```{r instmodis}
library(MODIStsp)
information <- MODIStsp_get_prodlayers("M*D13Q1")

information[3]
```


---

# Downloading the data

- Now, we send a long query to the MODIS database: 

```{r download_satel, eval=FALSE}
source('password.R')
MODIStsp(
  gui = FALSE,
  out_folder = "./",
  out_folder_mod = "./",
  selprod = "Vegetation_Indexes_16Days_1Km (M*D13A2)",
  bandsel = "NDVI",
  user = "basm92",
  password = password_here,
  start_date = "2020.06.01",
  end_date = "2020.06.01",
  verbose = FALSE,
  spatmeth = "file",
  spafile = "gadm/netherlands.shp",
  out_format = "GTiff"
)

```

---

# Import the data in R

- ..And we import the data in R:

```{r cropraster}
library(raster); library(terra)
# Import the file 
ndvi_raster <- raster('./MYD13A2.A2020153.h18v03.061.2020336102901.hdf')
# Set the CRS of the Netherlands to this file
crs_raster <- st_crs(ndvi_raster)
netherlands_transformed <- st_transform(netherlands, crs= crs_raster)
# Isolate only the relevant part overlapping the netherlands
ndvi_raster <- raster::mask(ndvi_raster, netherlands_transformed)
# Crop it
ndvi_raster <- terra::crop(ndvi_raster, netherlands_transformed)
```

```{r alternative_stuff, include=F, eval=F}
# Projecting ndvi to netherlands instead of the other way around
ndvi_raster <- raster('./MYD13A2.A2020153.h18v03.061.2020336102901.hdf')
ndvi_raster <- projectRaster(ndvi_raster, crs="WGS84")
ndvi_raster <- raster::mask(ndvi_raster, netherlands)
ndvi_raster <- terra::crop(ndvi_raster, netherlands)
```

---

# Plot the output

- Our file looks like this:

```{r pltveg, fig.width=8, fig.height=5.5, fig.align='center'}
plot(ndvi_raster)
```


---

# Main task

- Now we can proceed to our main task: computing average vegetation per polygon in our `netherlands` shapefile
  - Or rather, let's do this by municipality!
  
```{r down_minnica}
# Import
netherlands_munip <- geodata::gadm("Netherlands", level=2, path="./") |> st_as_sf()

# Set the CRS
netherlands_munip <- st_transform(netherlands_munip, crs = crs_raster)
```

- This can be done *very easily* through `raster`'s `extract` function:
  - This extracts a list of values from the raster for each polygon in the sf data frame:
  - Followed by some data wrangling:

```{r aggvgimun, cache=T}
values <- extract(ndvi_raster, netherlands_munip)
netherlands_munip <- netherlands_munip |>
  mutate(mean_ndvi = map_dbl(values, mean, na.rm=T))
```

---

# Inspect our output

- Finally, we can plot this:
  - Some of the highly urbanized areas have a low vegetation index and some of the more rural areas have a higher vegetation index


```{r pltveghoi, fig.align='center', fig.width=6, fig.height=4}
netherlands_munip |> 
  ggplot(aes(fill=mean_ndvi)) + 
  geom_sf() +
  scale_fill_viridis_c(direction = -1, option = 'turbo')
```


---


class: inverse, center, middle

# Part 4: Aggregate Vector to Raster


---

# Download Map to Start With

- We first download the map of the Netherlands through the `giscoR` package
  
```{r get_municipalities, cache=TRUE}
netherlands <- giscoR::gisco_get_lau(year='2020', country='Netherlands')

netherlands |> head(5)
```


---

# Plot The Map

- Let us have a look at `POP_DENS_2020`, the population density in 2020
  - This is but one of the variables in this spatial data.frame
  
```{r plot_nl, fig.align='center', fig.height=6, fig.width=6}
netherlands |> 
  ggplot(aes(fill=log(POP_DENS_2020))) +
  geom_sf()
```

---

# Make Raster of The Map

- Next, we make a grid dividing the Netherlands into areas of $x \times y$ latitude x longitude
  - Say $0.05 \times 0.05$ in this case
  
- The grid literally consists of pieces of area of 0.05 latitude by 0.05 longitude
  - This can be the unit of our analysis

```{r make_grid}
library(sf); library(stars); library(starsExtra)

grid <- st_make_grid(netherlands, square=T, cellsize=c(0.05, 0.05))
```


---

# Aggregate Original Vector Data to Raster

- Let us now use `raster::aggregate` to `aggregate` the variables of `netherlands` to the grid-level

- We can choose an _aggregation function_: this time, we use the mean
  - That is, for each box in the grid, we compute the corresponding `mean` values of the polygons and compute a geographical "weighted average" of our population density and other variables

```{r aggregate_grid}
per_grid <- raster::aggregate(netherlands, grid, FUN=mean)
```

---

# Look at Plot

- Let us look at the output:

```{r plot_gridded, fig.height=5, fig.align='center'}
per_grid |> 
  st_as_sf() |> 
  ggplot(aes(fill=POP_DENS_2020)) + geom_sf()
```


---

class: inverse, center, middle

# Extra: `geodata`

---

# The `geodata` package

- The `geodata` package is a package through which you can get all kinds of open-sourced georeferenced data

- Install it and load it through `pacman`:

```{r geodataload}
p_load(geodata)
```

- For example, geographical data can be downloaded through:

```{r download_nl}
nl <- geodata::gadm("Netherlands", level = 1, path = "./") |> st_as_sf()
```

---

# The `geodata` package

- You can also download other kinds of (raster and vector) data:
  - This can again be aggregate to our raster or vector data!

```{r download_temp, fig.height=4, fig.align='center'}
temperature <- geodata::worldclim_country("Netherlands",var = "tavg", path = "./")
temperature$NLD_wc2.1_30s_tavg_1 |> plot()
```


```{r eval=F, include=F}
netherlands_raster <- st_rasterize(netherlands |> 
                                     dplyr::select(POP_DENS_2020)
                                   )

coords <- st_coordinates(netherlands_raster) |> 
  summarize(xmin = min(x), xmax = max(x),
            ymin = min(y), ymax = max(y))
```

