---
title: "Historical Persistence & Geospatial Economics"
subtitle: "Applied Economics Research Course"
author: "Bas Machielsen"
aspectratio: 169
format: beamer
header-includes:
  - \usepackage{underscore}
---

```{r setup}
#| include: false
library(tidyverse); library(kableExtra); library(tinytable)
```

# Introduction

## Institutions

- Economic or other outcomes are not only influenced directly by relative prices, costs and benefits.
- They are also influenced by more latent, long-term factors, often called institutions.
  - Example: Acemoglu, Robinson and Johnson (2001):

  > (..) estimate the effect of institutions on economic performance. Europeans adopted very different colonization policies in different colonies, with different associated institutions. In places where Europeans faced high mortality rates, they could not settle and were more likely to set up extractive institutions. (..) Exploiting differences in European mortality rates as an instrument for current institutions, we estimate large effects of institutions on income per capita.

## Institutions

- Here's a world map with GDP/capita in 1995 for a selected number of countries:

```{r download_map}
#| fig-height: 4
#| fig-width: 8
#| fig-align: 'center'
#| warning: false
#| message: false
#| output-location: slide
library(sf); library(rnaturalearthdata); library(ivdoctr)
world_map <- rnaturalearthdata::countries50 |> st_as_sf()
wm <- world_map |> left_join(ivdoctr::colonial,
                             by = c("iso_a3" = "shortnam"))
wm |> ggplot(aes(fill=logpgp95)) + geom_sf()
```

## Institutions

- In this paper, Acemoglu, Johnson and Robinson (2001) take institutions to be _expropriation risk_: Measures risk of government appropriation of foreign private investment on a scale from 0 (least risk) to 10 (most risk), averaged over all years from 1985-1995.
- If one wants to find the relationship between institutions and economic well-being, one runs into the problem of _endogenity_:
  - Institutions causally affects economic growth
  - But economic growth also affects institutions!
  - Formally:
      - $\text{GDP/Cap}_i = \alpha_0 + \alpha_1 \text{Institutions}_i + \epsilon^1_i$
      - $\text{Institutions}_i = \beta_0 + \beta_1 \text{GDP/Cap}_i + \epsilon^2_i$
- If this is the _true_ data-generating process, an OLS regression will not yield the correct coefficient you are interested in ( $\alpha_1$ ).

## Instrumental Variables

- People have devised various strategies to solve this problem. The most often-used strategy is to use an _instrumental variable_
  - An instrumental variable is a variable that exogenously causes a change in $X$ while not directly affecting $Y$.
  - In the AJR (2001) case: $\text{Institutions}_i = \beta_0 + \beta_1 \text{GDP/Cap}_i + \underbrace{\beta_2 \text{SettlerMortality}_i}_{\text{Instrumental Variable}} + \epsilon^2_i$
  - Now, GDP per capita and Institutions are _endogenously_ determined as a result of _exogenous_ settler mortality

## Instrumental Variables

- You can rearrange these two equations to find that:

$$
\text{Institutions}_i = \dots + \underbrace{\left( \frac{\beta_1 \alpha_1 \beta_2}{1-\alpha_1 \beta_1} + \beta_2 \right)}_{= \frac{\beta_2}{1-\alpha_1\beta_1}} \cdot \text{SettlerMortality}_i + \dots
$$

$$
\text{GDP/Cap}_i = \dots + \left( \frac{\alpha_1 \beta_2}{1-\alpha_1 \beta_1} \right) \cdot \text{SettlerMortality}_i + \dots
$$

- Now you can find (_identify_) $\alpha_1$ by dividing $\text{Cov(GDP/Cap, SettlerMort)}$ by $\text{Cov(SettlerMort, Institutions)}$
- The key (and untestable) assumption here is that settler mortality does not directly influence GDP per capita in 1995.
- Loosely speaking, settler mortality represents an exogenous shock to institutions, allowing us to identify the influence of a change in institutions on GDP per capita

# Historical Persistence

## Back to historical persistence

- In general, we are dealing with outcomes that are determined endogenously: outcomes now and outcomes in the past are determined by other, latent factors
- To identify the influence of these latent factors, economists often use large-scale, influential events that shock these institutions. They compare outcomes in places that were initially similar, but some of them have been coincidentally exposed to certain events, whereas others have not.
- Usually, this research is based on spatial regression discontinuity designs, comparing places at one side of the border with places at the other side of the border (e.g. Dell, 2010, Lowes and Montero, 2021).

## Historical Persistence

- The set-up in general is:
  - Outcome $Y_i$ has "deep roots" caused by some $D_i$ long ago
  - This $D_i$ usually did not came to be randomly, but at the margin near some threshold, it did
  - Example (Dell 2011): Spanish conquest of Latin America was bounded by mountainous areas
- Comparing villages just within vs. just outside of the old imperial borders
- Other seminal papers:
  - Acemoglu et al. (2011), Occupied areas by Napoleon in Germany do better
  - Dell & Olken (2020): Positive development effects of extractive colonialism in Indonesia
  - Voigtlander & Voth (2012): Deep roots of persecution in Nazi Germany

## Spatial Regression Discontinuity

- Consider the example of **Roman roads**. Let's suppose that Roman roads, by influencing trade networks, stimulate development:

  - $Y_i = \alpha_0 + \alpha_1 \text{RomanRoad}_i + \alpha_2 \text{Environment}_i + \epsilon^1_i$

- But Roman Roads have not been built randomly by the Romans, but might have built in areas with more suitable environments for development:

  - $\mathbb{P}[\text{RomanRoad}_i] = \beta_0 + \beta_1 \text{Environment}_i + \epsilon^2_i$

- Then, comparing areas with roads to areas without roads doesn't work due to differences in the environment:

  - $\mathbb{E}[Y | RR = 1] = \alpha_0 + \alpha_1 + \alpha_2 \mathbb{E} [\text{Environment|RR = 1]}$
  - $\mathbb{E}[Y | RR = 0] = \alpha_0 + \alpha_2 \mathbb{E} [\text{Environment|RR = 0]}$

## Identification Strategy

- If we can find two places with the same expected value of getting a road, but only _one_ of them gets a road..
- According to the equations on the previous slide, this means two places with the same environment (take the expected value of $Y_i$ conditional on Environment to see that)
- Then, we can estimate $\alpha_1$ by:
  - $\mathbb{E}[Y | \text{RomanRoad \& P[RomanRoad]}]$ - $\mathbb{E}[Y | \text{NoRomanRoad \& P[RomanRoad]}]$
- This is what we attempt to do in a spatial regression discontinuity design: compare two arguably identical places with the same probability of treatment, where only one of them actually gets the treatment

# The Data

## Data

- I have two a couple of data sets on offer to you:
  - A dataset of French and German municipalities overlaid with Roman roads (`france_germany.shp`)
  - This dataset contains the distance of each municipality to a Roman road, as well as an indicator, whether a road streches through a municipality or not
  - A dataset of Dutch municipalities overlaid with the border of the former Roman empire (`netherlands_rome.geojson`)
  - Again including a variable for distance to the border, and an indicator of whether a present-day municipality was inside or outside the former Roman empire
- You are also free to use or compile a custom dataset using different settings and countries.

# Demonstration

## Demonstration

```{r demonstration}
#| fig-height: 3
#| fig-width: 3
#| fig-align: 'center'
#| cache: true
#| output-location: slide
library(cawd); library(giscoR)
roads <- cawd::darmc.roman.roads.major.sp |> st_as_sf()
lau <- giscoR::gisco_get_lau(year = '2019', country = c("France"))

xmin <- -10; xmax <- 20; ymin <- 40; ymax <- 80
box <- st_bbox(c(xmin=xmin, xmax=xmax, ymax=ymax, ymin=ymin), crs = st_crs(4326))
square <- st_sf(st_as_sfc(box))
fr <- st_intersection(lau, square)

fr |> ggplot() + geom_sf()
```

## What does this data.frame look like? {shrink=20}

```{r show_head}
#| output-location: slide
#| code-overflow: scroll
#| 
fr |> head(5)
```

## Merge this with the roads

```{r merge}
#| cache: true
numbers <- st_intersects(roads, fr)|>
  as.data.frame() |> select(row.id) |> pull() |> unique()

roads_in_fr <- roads |> filter(is.element(row_number(), numbers))
```

```{r plt}
#| fig-height: 3
#| fig-width: 3
#| fig-align: 'center'
#| cache: true
#| output-location: slide
ggplot() +
  geom_sf(data = fr) +
  geom_sf(data = roads_in_fr, color = 'blue')
```

## Compute the distance

```{r mindist}
#| echo: true
#| cache: true
minimum_distances <- fr |>
  st_centroid() |>
  st_distance(roads_in_fr) |>
  apply(1, min)
```

```{r show_vect}
minimum_distances[1:10]
```

```{r add_to_df}
fr <- fr |>
  mutate(min_dist_to_road = minimum_distances)
```

## Plot the result

```{r plot_rds}
#| fig-height: 3
#| fig-width: 6
#| fig-align: 'center'
#| cache: true
#| output-location: slide
ggplot() +
  geom_sf(data = fr, aes(fill = min_dist_to_road), size = 0.0001) +
  scale_fill_viridis_c() +
  geom_sf(data = roads_in_fr, color = 'orange')
```

## A small analysis

```{r analysis}
#| output-location: slide
library(fixest); library(modelsummary)
model1 <- fixest::feols(POP_DENS_2019 ~ min_dist_to_road, data = fr)

modelsummary(list("Pop. Density"=model1), gof_map = c("r.squared", "nobs"), stars=T) |>
  theme_latex(
    resize_width = 0.4,
    resize_direction = 'both'
  )
```

## Your task

- Find municipality-level or geographical outcomes that suit an *interesting research question*
  - Present-day or more historical/macro-level outcomes
- Make sure you control for environments!
  - Overlapping regions (provinces, supersets of municipalities)
  - But also possible: terrain, other initial conditions (coastal municipalities)
  - Make it plausible that you compare apples with apples
  - Some roads have also been built with a destination in mind: how to control for that?
- Explore mechanisms through which the effect can work

# Geospatial Economics

## Introduction

- In economics, many research questions can be answered using geospatial data. For example, questions related to economic development (Beyer et al., 2021; Besley et al., 2022) make frequent use of variation between different geographic units and compare them in aspects such as nightlight density and electricity consumption.
- Similarly, questions related to environmental economics can also be answered using similar data on the basis of geospatial variation (Castells-Quintana et al., 2021; Felbermayer et al., 2022).
- This theme will focus on similar research questions with the aim of introducing students to geospatial data wrangling and econometric methods.

## Example

- This theme is a little bit broader than the Historical Persistence theme
- Many research questions are possible.
- Usually, your outcome variable is based on a *shapefile* containing an interesting outcome, e.g. nightlights, species distribution, pollution
- As an example, I have prepared a dataset and analysis of the following research question:
  - Does the air quality influence the performance of French schools?

## Loading the Data

- Load the school performance dataset (source: [data.gouv.fr](https://www.data.gouv.fr/fr/datasets/indicateurs-de-valeur-ajoutee-des-colleges/#/resources))

```{r load_sp}
#| echo: true
library(here)
school_performance <- read_delim(here('data', 'schools', 'fr-en-indicateurs-valeur-ajoutee-colleges.csv'))
```

- Filter so that the year is 2022:

```{r filt_sp}
#| echo: true
school_performance <- school_performance |>
  filter(Session == 2022)
```

## Geocoding the Schools

- We have to geocode the schools using a dataset containing the coordinates of these schools:

```{r coords}
#| echo: true
coordinates <- read_delim(here('data', 'schools', 'fr-en-adresse-et-geolocalisation-etablissements-premier-et-second-degre.csv')) |>
  select(numero_uai, latitude, longitude) |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  st_as_sf(coords=c('longitude', 'latitude'), crs=4326)
```

## Filtering

- Filter only Metropolitan France (for the sake of visualization)

```{r plot}
#| fig-height: 2.8
#| fig-width: 4
#| fig-align: 'center'
#| output-location: slide
library(geodata)
france <- geodata::gadm("France", path=tempdir()) |>
  st_as_sf()
coordinates <- coordinates |>
  st_filter(france)

coordinates |>
  ggplot() + geom_sf(size=0.1)
```

## Importing Pollution Data

- This is annual NOx concentration data (Source: [European Environment Agency](https://www.eea.europa.eu/en/datahub/datahubitem-view/82700fbd-2953-467b-be0a-78a520c3a7ef?activeAccordion=1092391%2C1092389)):
  - NOx stands for Nitrogen Oxides—a group of harmful gases (mainly NO and NO₂) produced when fossil fuels (like gasoline, diesel, coal) burn at high temperatures. Common sources are cars & trucks, power plants and factories
- Read the file, downsample it (to keep it computationally manageable)
- Put it in the same CRS, keep only raster squares that are in France:

```{r im_pol}
#| echo: true
library(stars); library(dplyr)
sf_use_s2(F)
nox <- read_stars(here('data', 'schools', 'nox_avg_22.tif'))
nox_lowres <- nox |> st_transform(st_crs(france))
nox_lowres <- st_warp(src=nox_lowres, cellsize = c(0.1, 0.1), crs=4326)
nox_lowres <- nox_lowres |> st_crop(st_union(france))
```

## Plot Result

```{r plotnox}
#| fig-height: 3
#| fig-width: 4.5
#| fig-align: 'center'
#| output-location: slide
ggplot() + geom_stars(data=nox_lowres) + scale_fill_viridis_c()
```

## Link Coordinates to Schools {shrink=20}

```{r lcts}
#| echo: true
geocoded_schools <- inner_join(coordinates, school_performance, by = c('numero_uai'='UAI'))
geocoded_schools |> head(5)
```

## Link Geocoded Schools to Pollution {shrink=20}

```{r lstp}
#| echo: true
pollution_values <- st_extract(nox_lowres, geocoded_schools, FUN=mean, na.rm=T)
geocoded_schools <- geocoded_schools |>
  mutate(pollution_values=pollution_values$nox_avg_22.tif)
geocoded_schools |> head(5)
```

## Small Analysis

- Clean up variable names, and run a regression

```{r mdl}
#| output-location: slide
library(fixest); library(janitor)
geocoded_schools <- geocoded_schools |>
  clean_names()

model <- feols(va_de_la_note_g ~ pollution_values + nb_mentions_tb_g, data=geocoded_schools)
modelsummary(list("Avg. Grade"=model), gof_map = c("r.squared", "nobs"), stars=T) |>
  theme_latex(
    resize_width = 0.4,
    resize_direction = 'both'
  )
```

# Logistics

## Course Schedule

- Upload research plan in Osiris 13 May 2025 at the latest
- Complete paper and upload in Osiris 26 June 2025 at the latest
- Lecture 1: Plenary Lecture, Introduction
- Lecture 2: Discussion of RQs (Send them on Teams!)
- Lecture 3 + 4: Plenary Lectures, Geospatial Data
  - Depending on time: room to deliberate
- Lectures 5-7: Feedback sessions, room to work and discuss
- Lecture 8: Presentations, feedback from supervisor and co-reader

## Expected output

- At the end of week 3 you will upload your research plan in Osiris, including problem statement, (preliminary) literature review and research approach. This plan and your performance in the research group so far will be the input for your supervisors' go/no go decision (crucial with respect to your compliance with the effort requirements).
- In the meeting in week 4, the focus is on the research plan, which is a crucial element in the overall research process. It should contain the motivation for the central problem statement and a plan how to do the actual research.
- Clearly, all members of the supervision group should receive each other's' draft research plan in time, that is, a few days before the meeting itself; to make the meeting as useful as possible, everyone should have carefully read the different research plans, should have given each of these a thought and should have prepared some questions or suggestions.

## Expected output

- In each presentation (meeting 4, meeting 6, meeting 8) session, every student gives a short and focused presentation of his/her research project – corresponding to the stage the project is in – and elaborates both on the contents and the used research method and methodology, on choices that have been made and on solved or unsolved problems they have encountered in their work.
- Each presentation is followed by a general discussion where all participants are supposed to contribute by asking questions for further explanation as well as critical questions about the actual research and by providing constructive suggestions and comments. _This feedback needs to be written down in a document and handed in at the start of each session_
- **Final Paper Due**: 26 June 2025

## Contact

- You can always contact me at [a.h.machielsen@uu.nl](mailto:a.h.machielsen@uu.nl)
- Also: a **Microsoft Teams group**
- Data available on course website.

## Data Sources

- World Values Survey: https://www.worldvaluessurvey.org - Geographically coded information about norms & values
- Demographics & Health Survey: https://dhsprogram.com/data/ - Also geocoded information about D&H
- [PRIO GRID Gridded Data](https://grid.prio.org/#/codebook)
- Eurostat: https://ec.europa.eu/eurostat/web/main/data/database - Database about European countries on aggregate (country) and more disaggregated levels
- [Waar Staat Je Gemeente?](https://www.waarstaatjegemeente.nl/): Dutch Municipality Data
- [Statistiques Locales (FR)](https://statistiques-locales.insee.fr/#c=home): French Municipality Data
- Global Climate Database: http://www.worldclim.org/
- My own website data overview (includes municipality data): [here](https://bas-m.netlify.app/blog/2022-12-06-overview-of-economic-history-datasets-and-databases/)
- Clio-Infra: https://github.com/basm92/Clio
- Standard scholarly measures of politics: https://github.com/xmarquez/democracyData
- Correlates of war, armed conflicts: https://rdrr.io/cran/peacesciencer/
- Political protests: https://acleddata.com/
- Labor conflicts: https://datasets.iisg.amsterdam/dataverse/labourconflicts
- World Bank, IMF, Countries' National Statistical Agencies, Countries' Parliaments
- [Rspatialdata](https://rspatialdata.github.io/)

## References & Literature

Acemoglu, D., Cantoni, D., Johnson, S., & Robinson, J. A. (2011). The consequences of radical reform: The French Revolution. American economic review, 101(7), 3286-3307.

Beach, B., & Hanlon, W. W. (2022). Culture and the historical fertility transition. The Review of Economic Studies.

Dell, M. (2010). The persistent effects of Peru's mining mita. Econometrica, 78(6), 1863-1903.

Dell, M., & Olken, B. A. (2020). The development effects of the extractive colonial economy: The dutch cultivation system in java. The Review of Economic Studies, 87(1), 164-203.

Lowes, S., & Montero, E. (2021). Concessions, violence, and indirect rule: evidence from the Congo Free State. The Quarterly Journal of Economics, 136(4), 2047-2091.

Lowes, S. (2022). Kinship Structure and the Family: Evidence from the Matrilineal Belt (No. w30509). National Bureau of Economic Research.

## References & Literature

Jones, B. F., & Olken, B. A. (2005). Do leaders matter? National leadership and growth since World War II. The Quarterly Journal of Economics, 120(3), 835-864.

Voth, H. J. (2021). Persistence–myth and mystery. In The handbook of historical economics (pp. 243-267). Academic Press.

Voigtländer, N., & Voth, H. J. (2012). Persecution perpetuated: the medieval origins of anti-Semitic violence in Nazi Germany. The Quarterly Journal of Economics, 127(3), 1339-1392.

Besley, T., Burgess, R., Khan, A., & Xu, G. (2022). Bureaucracy and development. Annual Review of Economics, 14(1), 397-424.

Beyer, R. C., Franco-Bedoya, S., & Galdo, V. (2021). Examining the economic impact of COVID-19 in India through daily electricity consumption and nighttime light intensity. World Development, 140, 105287.

## References & Literature

Donaldson, D., & Storeygard, A. (2016). The view from above: Applications of satellite data in economics. Journal of Economic Perspectives, 30(4), 171-198.

Castells-Quintana, D., Dienesch, E., & Krause, M. (2021). Air pollution in an urban world: A global view on density, cities and emissions. Ecological Economics, 189, 107153.

Felbermayr, G., Gröschl, J., Sanders, M., Schippers, V., & Steinwachs, T. (2022). The economic impact of weather anomalies. World Development, 151, 105745.

Beach, B., & Hanlon, W. W. (2018). Coal smoke and mortality in an early industrial economy. The Economic Journal, 128(615), 2652-2675.

Hanlon, W. W. (2020). Coal smoke, city growth, and the costs of the industrial revolution. The Economic Journal, 130(626), 462-488.
