---
title: "Bounding Box to Polygon via Geocoding"
format: html
---

## Introduction

Sometimes we have a list of place names (cities, regions, countries) and want to turn them into spatial objects we can map or analyze. One way to do this is by **geocoding**: looking up the geographic coordinates of a place by name.

This notebook shows how to:

1. Look up the **bounding box** of a location (a rectangle covering the area) using OpenStreetMap data.
2. Convert that bounding box into a **polygon** (an `sf` geometry).
3. Apply this process to **every row** of a data frame.

## Step 1: Load packages

We use two packages:

- `osmdata` — to query OpenStreetMap and retrieve bounding boxes by place name.
- `sf` — to work with spatial (vector) data in R.

```{r}
library(osmdata)
library(sf)
```

## Step 2: Write a function to convert a location name to a polygon

`getbb()` from `osmdata` takes a place name like `"Utrecht"` and returns a 2×2 matrix with the minimum and maximum longitude (`x`) and latitude (`y`) of that place.

We then use those four corner coordinates to build a rectangular polygon with `sf`.

```{r}
bbox_to_polygon <- function(location) {
  # Look up the bounding box from OpenStreetMap
  bbox <- getbb(location)

  # Pull out the four corners
  min_x <- bbox["x", "min"]
  max_x <- bbox["x", "max"]
  min_y <- bbox["y", "min"]
  max_y <- bbox["y", "max"]

  # Build a coordinate matrix tracing the rectangle (5 points: close the ring)
  coords <- matrix(c(
    min_x, min_y,
    max_x, min_y,
    max_x, max_y,
    min_x, max_y,
    min_x, min_y   # back to start to close the polygon
  ), ncol = 2, byrow = TRUE)

  # Create an sf polygon in WGS 84 (the standard GPS coordinate system)
  polygon <- st_polygon(list(coords))
  sf_polygon <- st_sfc(polygon, crs = 4326)

  return(sf_polygon)
}
```

## Step 3: Try it for a single location

Let's test the function on Utrecht to make sure it works before applying it to a whole data frame.

```{r}
polygon <- bbox_to_polygon("Utrecht")
polygon
```

## Step 4: Apply the function to every row of a data frame

Suppose we have a data frame with a column called `location` containing place names. We want to add a column `polygon` with the corresponding spatial polygon for each row.

We write a small helper that extracts the location from a single row, then use `rowwise()` + `mutate()` to apply it across all rows.

```{r}
#| eval: false

# Helper: geocode one row
polygon_from_row <- function(row) {
  loc <- row$location
  bbox_to_polygon(loc)
}

# Apply to all rows of df (replace df with your actual data frame)
df <- df |>
  rowwise() |>
  mutate(polygon = polygon_from_row(cur_data()))
```

> **Note:** `rowwise()` makes `mutate()` operate one row at a time, and `cur_data()` gives access to the current row as a mini data frame. Each call to `polygon_from_row()` makes a live request to OpenStreetMap, so this may be slow for large data frames.
