---
title: "Building a Grid Around the Roman Empire's Northern Border"
format: html
execute:
  freeze: auto
---

## Introduction

This notebook constructs a **raster grid** covering the area around the northern border of the Roman Empire. The goal is to create a spatial dataset that can be used in a regression discontinuity design (RDD): each grid cell gets a distance to the Roman border and an indicator for whether it falls inside or outside the empire.

Working with a regular grid (rather than administrative units like municipalities) is useful because it avoids arbitrary boundaries and allows for a uniform spatial analysis across countries.

We will:

1. Load the map of Europe and the Roman Empire boundary.
2. Extract the northern border of the Roman Empire.
3. Remove a small section of coastline that is not relevant.
4. Build a 200 km buffer around the northern border.
5. Clip that buffer to the European land area.
6. Create a regular grid within this area.
7. Compute each cell's distance to the border and its inside/outside status.
8. Build the signed running variable and visualise the result.

## Step 1: Load packages and spatial data

```{r}
library(cawd)             # historical boundary data (Roman Empire)
library(rnaturalearthdata) # country polygons
library(tidyverse)
library(sf)
library(nngeo)            # nearest-neighbour geometry tools
# remotes::install_github("michaeldorman/nngeo")  # if not yet installed
```

We load a map of Europe from Natural Earth and the Roman Empire polygon from `cawd`.

```{r}
# Map of Europe (sovereign states)
europe <- rnaturalearthdata::sovereignty50 |>
  st_as_sf() |>
  filter(continent == "Europe")

# Roman Empire at 117 AD
re <- cawd::awmc.roman.empire.117.sp |> st_as_sf()
```

## Step 2: Extract the northern border of the Roman Empire

`st_boundary()` converts the Roman Empire polygon into its outer border (a line). We then clip that line to only the countries that contain the **northern** frontier — from the Netherlands in the west to Bulgaria in the east.

```{r}
# Full outer boundary of the Roman Empire
boundary_re <- st_boundary(re)

# Countries along the northern frontier
countries_with_north_border <- europe |>
  filter(is.element(sovereignt, c(
    "Netherlands", "Germany", "Croatia",
    "Austria", "Hungary", "Slovakia",
    "Ukraine", "Romania", "Bulgaria",
    "Republic of Serbia"
  )))

# Keep only the border segments that pass through these countries
northernmost_lines <- st_intersection(boundary_re, countries_with_north_border)
```

## Step 3: Remove the southern Croatian coastline

The intersection above accidentally picks up part of Croatia's Adriatic coast, which is a southern border — not the northern frontier we want. We define a small polygon covering that area and subtract it.

```{r}
coords <- matrix(c(
  15, 46,
  20, 42,
  10, 42,
  10, 46,
  15, 46
), ncol = 2, byrow = TRUE)

polygon <- st_polygon(list(coords)) |>
  st_sfc(crs = 4326)

northernmost_border <- northernmost_lines |>
  st_difference(polygon)
```

## Step 4: Build a 200 km buffer around the northern border

We want to study areas *near* the Roman border — within 200 km on either side. `st_buffer()` creates a band of that width around the border line. We temporarily enable S2 geometry (spherical) for this step, as it is more accurate for wide buffers.

```{r}
sf_use_s2(TRUE)
buffer_around_northern_border <- st_buffer(northernmost_border, 200000)  # metres
```

## Step 5: Clip the buffer to European land

The buffer extends over sea areas that we do not need. We intersect it with the European land polygons to keep only the terrestrial area around the border.

```{r}
sf_use_s2(FALSE)
area_around_re_border <- europe |>
  st_intersection(buffer_around_northern_border)

# Quick visual check
area_around_re_border |>
  ggplot() + geom_sf()
```

## Step 6: Build a regular grid over this area

`st_make_grid()` lays a 100×100 regular grid over the bounding box of the study area. We then keep only cells that fall *within* the actual land area (discarding cells over sea or outside our region entirely).

```{r}
grid <- st_make_grid(area_around_re_border, n = 100)

# Keep only cells that fall within the study area
filter <- st_within(grid, st_union(area_around_re_border)) |> lengths() > 0
final_grid <- grid[filter]
```

## Step 7: Convert to a raster and save

We convert the grid to a `stars` raster object (each cell gets a value of 1) and save it as a GeoTIFF for later use.

```{r}
library(stars)
library(starsExtra)

raster_grid <- final_grid |> stars::st_as_stars()
final_raster_grid <- raster_grid[raster_grid["values"] == 1]

stars::write_stars(final_raster_grid, "../example_code/raster_re.tiff")
```

A quick plot to check the result alongside the full Roman Empire boundary:

```{r}
ggplot() +
  geom_stars(data = final_raster_grid) +
  geom_sf(data = boundary_re)
```

## Step 8: Compute distance to the border for each cell

`starsExtra::dist_to_nearest()` computes the distance from each raster cell centre to the nearest segment of the northern border line. The result is stored as a new `distance` attribute on the raster.

```{r}
nearest_feat <- starsExtra::dist_to_nearest(final_raster_grid, northernmost_lines)
final_raster_grid$distance <- nearest_feat$d
```

## Step 9: Flag cells as inside or outside the Roman Empire

We convert the raster cell centres to points and use `st_within()` to check which ones lie inside the Roman Empire polygon. The result is a logical vector that we store as `inside_re` (1 = inside, 0 = outside, NA = not a valid cell).

```{r}
grid_sf <- st_as_sfc(final_raster_grid, as_points = TRUE)
logical <- as.logical(st_within(grid_sf, re))

final_raster_grid <- final_raster_grid |>
  mutate(inside_re = case_when(
    !is.na(logical) & values == 1 ~  1,
     is.na(logical) & values == 1 ~  0,
    TRUE                          ~ NA
  ))
```

## Step 10: Build the signed running variable and visualise

As in a standard RDD, the running variable is the distance to the border, made **positive** for cells inside the Roman Empire and **negative** for those outside.

```{r}
final_raster_grid <- final_raster_grid |>
  mutate(running = if_else(inside_re == 1, distance, -distance))
```

```{r}
# Inside/outside map
final_raster_grid["inside_re"] |> plot()

# Distance map
final_raster_grid["distance"] |> plot()

# Running variable map
ggplot() +
  geom_stars(data = final_raster_grid["running"]) +
  scale_fill_viridis_c()
```
