---
title: "Intersecting the Roman Empire with Modern European Countries"
format: html
---

## Introduction

The Roman Empire covered large parts of what is now Europe, North Africa, and the Middle East. For many research questions in historical economics we only care about the **European** portion — for example, to study long-run persistence of Roman institutions within present-day European countries.

This notebook shows how to spatially intersect the Roman Empire polygon with modern European country boundaries to obtain exactly that overlap. We also take a quick look at the ORBIS road network nodes that `cawd` provides.

We will:

1. Load the Roman Empire polygon and a map of European countries.
2. Compute their spatial intersection.
3. Map the result.

## Step 1: Load packages and spatial data

```{r}
library(sf)
library(tidyverse)
library(rnaturalearth)
library(cawd)
```

We load two spatial datasets:

- **Roman Empire** — the `cawd` package contains `awmc.roman.empire.117.sp`, the boundary of the Roman Empire at 117 AD (its greatest extent). We convert it to an `sf` object.
- **European countries** — `rnaturalearthdata::countries50` provides world country polygons at 1:50m resolution. We filter to Europe only.

```{r}
re  <- cawd::awmc.roman.empire.117.sp |> st_as_sf()

eur <- rnaturalearthdata::countries50 |>
  st_as_sf() |>
  filter(continent == "Europe")
```

## Step 2: Compute the intersection

`st_intersection()` returns the parts of European countries that overlap with the Roman Empire — in other words, the portions of modern Europe that were once under Roman control.

Before running the intersection we disable S2 geometry (`sf_use_s2(FALSE)`). S2 uses a spherical Earth model, which can cause problems when polygons have small gaps or imprecise edges (as historical boundaries often do). Switching to flat (planar) geometry avoids these errors.

```{r}
sf::sf_use_s2(FALSE)

eur_in_re <- st_intersection(re, eur)
```

## Step 3: Map the result

Each feature in `eur_in_re` is the part of a European country that fell within the Roman Empire. The map shows at a glance which countries were (partially or fully) included.

```{r}
eur_in_re |>
  ggplot() +
  geom_sf()
```

