---
title: "Road Infrastructure as a Share of Municipal Area: France"
format: html
execute:
  freeze: auto
---

## Introduction

Road density is a common proxy for infrastructure in historical persistence research. A municipality with more roads relative to its area is better connected — and that connectivity may reflect centuries of investment or geographic advantage.

This notebook computes the **share of each French municipality's area covered by roads**. The approach is:

1. Draw a 10-metre buffer around each road segment (approximating its physical footprint).
2. Intersect those buffers with municipality boundaries.
3. Sum the buffered road area within each municipality and divide by total municipal area.

The result is a variable between 0 and 1 that can be merged into any France-level dataset for use as a control or outcome.

**Data sources:**

- Road network: IGN ROUTE500 (France, 2021) — downloaded directly from IGN.
- Municipality boundaries: GADM level-5 administrative boundaries for France.

## Step 1: Load packages

```{r}
library(sf)
library(tidyverse)
library(archive)    # for extracting .7z files
library(geodata)    # for downloading GADM boundaries
```

## Step 2: Set up a temporary directory

We download raw data into a temporary directory so nothing is written permanently to the project folder.

```{r}
tmp <- tempdir()
```

## Step 3: Download and extract the IGN roads data

The IGN distributes ROUTE500 as a `.7z` archive. We download it with `download.file()` and extract it using the `archive` package, which supports the 7z format.

This file is large (~200 MB) so the download may take a few minutes.

```{r}
roads_url  <- "https://data.geopf.fr/telechargement/download/ROUTE500/ROUTE500_3-0__SHP_LAMB93_FXX_2021-11-03/ROUTE500_3-0__SHP_LAMB93_FXX_2021-11-03.7z"
roads_7z   <- file.path(tmp, "ROUTE500.7z")

download.file(roads_url, destfile = roads_7z, mode = "wb")
archive::archive_extract(roads_7z, dir = tmp)
```

## Step 4: Import the road network

The shapefile we need is `TRONCON_ROUTE.shp` — each row is one road segment. We filter out segments without a route number (`NUM_ROUTE`), which removes non-classified tracks. To keep memory use manageable we work with the first 100,000 segments.

```{r}
shp_path <- file.path(
  tmp,
  "ROUTE500_3-0__SHP_LAMB93_FXX_2021-11-03",
  "ROUTE500", "1_DONNEES_LIVRAISON_2022-01-00175",
  "R500_3-0_SHP_LAMB93_FXX-ED211", "RESEAU_ROUTIER",
  "TRONCON_ROUTE.shp"
)

france_routes <- st_read(shp_path) |>
  filter(!is.na(NUM_ROUTE))

france_routes <- france_routes[1:100000, ]

ggplot(france_routes) + geom_sf()

# Save the CRS — we will reuse it later
crs <- st_crs(france_routes)
```

## Step 5: Download municipality boundaries from `giscoR`

`geodata::gadm()` downloads administrative boundaries directly from the GISCO database. This gives us French communes (the finest level). We reproject to match the roads CRS and compute each commune's total area in m².

```{r}
library(giscoR)
france_communes <- giscoR::gisco_get_lau(country="France") |>
  st_transform(crs) |>
  rename(total_area = AREA_KM2)
```

## Step 6: Buffer the roads

`st_buffer()` requires a projected CRS in metres to produce a meaningful buffer distance. We:

1. Temporarily reproject the roads to UTM zone 30 (metres).
2. Draw a 10-metre buffer around each segment — approximating its physical road surface.
3. Reproject the buffers back to the original CRS so they align with the communes.

```{r}
france_routes <- st_transform(
  france_routes,
  "+proj=utm +zone=30 +datum=WGS84 +units=m +no_defs"
)

roads_buffer <- st_buffer(france_routes, 10)
roads_buffer <- roads_buffer |> st_transform(crs)
```

## Step 7: Intersect road buffers with communes

`st_intersection()` clips each road buffer to the commune it falls in. The result is a data frame where each row is the piece of a road buffer that lies within a specific commune. We then compute the area of each piece.

```{r}
intersected       <- st_intersection(roads_buffer, france_communes)
intersected$area  <- st_area(intersected)
```

## Step 8: Aggregate road area by commune

We sum the buffered road area within each commune (`GID_5` is the unique commune identifier), then join the totals back onto the commune data frame.

The road surface share (`sufrace`) is then capped at 1 (in case rounding or overlapping buffers push it slightly above) and set to 0 for communes with no roads in this subset.

```{r}
road_area_by_commune <- aggregate(
  intersected$area,
  by  = list(intersected$GID_5),
  FUN = sum
)

france_communes <- france_communes |>
  left_join(road_area_by_commune, by = c("GID_5" = "Group.1")) |>
  mutate(sufrace = as.numeric(x) / total_area) |>
  mutate(sufrace = case_when(
    sufrace > 1  ~ 1,
    is.na(sufrace) ~ 0,
    TRUE           ~ sufrace
  ))
```

## Step 9: Export

We save the enriched commune dataset as a shapefile. Note that this file only reflects the **first 100,000 road segments**. To get complete coverage, run the same workflow on the full road network and sum the `sufrace` columns across all batches.

```{r}
#| eval: false
write_sf(france_communes, file.path(tmp, "france_communes_first_100000.shp"))
```

## Step 10: Map the result

We map the log-transformed road surface share. `log(1 + sufrace)` compresses the right tail so that the colour scale is not dominated by a few highly urbanised communes.

```{r}
france_communes |>
  ggplot(aes(fill = log(1 + sufrace))) +
  geom_sf(lwd = 0.0001) +
  scale_fill_viridis_c() +
  labs(fill = "log(1 + road share)")
```
