---
title: "Regression Discontinuity: Roman Empire Border and Population Density in the Netherlands"
format: html
---

## Introduction

Does historical inclusion in the Roman Empire have a lasting effect on economic development today? One way to study this is with a **regression discontinuity design (RDD)**: municipalities just inside the old Roman border are compared to those just outside it. Because the border was drawn for historical/military reasons rather than based on local economic conditions, crossing it is essentially random — making it a valid quasi-experiment.

In this notebook we use Dutch municipality data to:

1. Load the spatial dataset and visualise the Roman Empire boundary.
2. Construct a signed running variable (distance to the border, positive inside, negative outside).
3. Merge in population data and compute population density.
4. Run and visualise the RDD.

## Step 1: Load packages and data

We use `sf` for spatial data and `tidyverse` for data manipulation. The dataset `netherlands_roman_updated.geojson` contains Dutch municipalities with two key variables already prepared: `is_in_re` (1/0) and `distance_to_border` (in metres, signed: positive inside the empire, negative outside).

```{r}
library(tidyverse)
library(sf)

ned <- st_read("./netherlands_roman_updated.geojson")
```

A first map to see which municipalities were inside the Roman Empire:

```{r}
ned |>
  ggplot(aes(fill = is_in_re)) +
  geom_sf()
```

## Step 2: Construct the running variable

The RDD requires a single continuous **running variable** that is centred on the cutoff (the Roman border, at zero). The `distance_to_border` column is already signed:

- **Positive** for municipalities *inside* the Roman Empire.
- **Negative** for municipalities *outside* it.

We rename it to `running` for clarity.

```{r}
data <- ned |>
  mutate(running = distance_to_border)
```

We can map the running variable to confirm it looks right — positive values should appear in the south of the Netherlands (inside the empire) and negative in the north:

```{r}
data |>
  ggplot(aes(fill = running)) +
  geom_sf()
```

## Step 3: Load population data and merge

We load a CSV file with 2022 population counts per municipality (sourced from CBS via `cbsodataR`).

We then join the population data to the spatial data frame by matching municipality names.

```{r}
population <- read_csv("./popned.csv")

data <- data |>
  left_join(population, by = c("NAME_2" = "gemeenten_2022"))
```

## Step 4: Compute population density

Raw population counts are hard to compare across municipalities of different sizes. We compute **population density** as inhabitants per km² by dividing total population by the municipality area (stored in `area`, which is in m²; multiplying by 1,000,000 converts to km²).

```{r}
data <- data |>
  mutate(density = bevolking_totaal / as.numeric(area) * 1000000)
```

A map of population density across the Netherlands:

```{r}
data |>
  ggplot(aes(fill = density)) +
  geom_sf()
```

## Step 5: Run the regression discontinuity

We use the `rdrobust` package, which implements the standard RDD estimator with data-driven bandwidth selection.

- `rdplot()` produces a visual of the discontinuity: it bins observations along the running variable and plots average density on either side of the cutoff (zero).
- `rdrobust()` estimates the size of the jump at the cutoff and reports the result with robust confidence intervals.

```{r}
library(rdrobust)

rdrobust::rdplot(y = data$density, x = data$running, c = 0)

rdrobust::rdrobust(y = data$density, x = data$running, c = 0) |>
  summary()
```

## Step 6: Scatter plot of the raw data

A simple scatter plot puts each municipality as a dot, with its running variable on the x-axis and its population density on the y-axis. The cutoff at zero separates the two groups. This gives a quick informal sense of whether a jump is visible at the border.

```{r}
data |>
  ggplot(aes(x = running, y = density)) +
  geom_point()
```
