---
title: "The Effect of Historical Treatment on Population Density: France-Germany Border Region"
format: html
execute: 
  freeze: auto
---

## Introduction

This notebook analyses the relationship between a historical treatment variable and present-day population density in the France-Germany border region. The dataset contains LAU-level regions straddling the French-German border, with a binary indicator (`treatment`) marking which regions received a Roman road, the historical treatment of interest.

To isolate the effect of the treatment from confounding regional characteristics, we estimate a series of OLS regressions with increasingly granular **fixed effects** at the NUTS-2 and NUTS-3 level. Fixed effects absorb any time-invariant differences between regions at those levels, so the treatment coefficient reflects variation *within* those groupings. The NUTS-2/3 identifiers are not part of the LAU dataset, so we retrieve them from Eurostat via the `giscoR` package and assign them to each LAU through a spatial join.

We will:

1. Load and map the spatial data.
2. Filter to regions with a valid treatment assignment.
3. Fetch NUTS-2 and NUTS-3 boundaries from `giscoR` and spatially join them to the LAU data.
4. Estimate three regression models with progressively stricter fixed effects.
5. Display the results in a publication-style table.

## Step 1: Load packages and data

```{r}
library(sf)
library(tidyverse)
library(giscoR)
```

We read in the France-Germany shapefile. Each row is a LAU region with spatial geometry, a treatment indicator (`treatment`), and population density (`POP_DENS_2016`).

```{r}
fr_gr <- sf::st_read("france_germany_updated.geojson")
```

A quick map to get a sense of the geography:

```{r}
fr_gr |>
  ggplot() +
  geom_sf()
```

## Step 2: Filter to regions with a treatment assignment

Some regions have `NA` for `treatment` — these are regions where treatment status is unknown or not applicable (e.g. regions far from the border). We drop them so that all three models are estimated on the same sample.

```{r}
data <- fr_gr |>
  filter(!is.na(treatment))

data |>
  ggplot(aes(fill = treatment)) + geom_sf()
```

## Step 3: Attach NUTS-2 and NUTS-3 identifiers via `giscoR`

The LAU data does not contain NUTS-2/3 codes. We fetch the NUTS-3 boundaries for France and Germany from Eurostat (year 2016, to match the population data) using `giscoR`, then use a spatial join to assign each LAU to its parent NUTS-3 region. Since NUTS-3 codes embed the NUTS-2 code as a prefix (e.g. NUTS-3 `FR101` → NUTS-2 `FR10`), we can derive the NUTS-2 identifier with a simple substring operation.

```{r}
sf::sf_use_s2(FALSE)

nuts3 <- giscoR::gisco_get_nuts(
  year       = "2016",
  nuts_level = 3,
  country    = c("FR", "DE"),
  resolution = "03"
) |>
  select(nuts_3 = NUTS_ID)

# Make sure CRS matches before joining
nuts3 <- st_transform(nuts3, st_crs(data))

data <- data |>
  st_join(nuts3, largest=TRUE) |>
  mutate(nuts_2 = substr(nuts_3, 1, 4))
```

A quick check: how many LAU regions received a NUTS-3 assignment?

```{r}
cat("Regions with NUTS-3 assigned:", sum(!is.na(data$nuts_3)), "\n")
cat("Regions missing NUTS-3:      ", sum( is.na(data$nuts_3)), "\n")
```

## Step 4: Estimate three regression models

We use `fixest::feols()`, which is a fast OLS estimator that supports high-dimensional fixed effects.

- **Model 1** — a simple bivariate regression: does treatment predict population density?
- **Model 2** — adds NUTS-2 fixed effects. This controls for broad regional differences (e.g. between Alsace and Baden-Württemberg), so we compare treated and untreated regions *within the same NUTS-2 area*.
- **Model 3** — adds both NUTS-2 and NUTS-3 fixed effects. This is the most restrictive specification, absorbing even finer regional variation.

The fixed effects are specified after the `|` in the formula.

```{r}
library(fixest)

m1 <- fixest::feols(POP_DENS_2016 ~ treatment,                    data = data)
m2 <- fixest::feols(POP_DENS_2016 ~ treatment | nuts_2,           data = data)
m3 <- fixest::feols(POP_DENS_2016 ~ treatment | nuts_2 + nuts_3,  data = data)
```

## Step 5: Display results in a regression table

`modelsummary()` puts all three models side by side. `stars = TRUE` adds the conventional significance stars (*, **, ***). The table makes it easy to see how the treatment coefficient changes as we add more controls.

```{r}
library(modelsummary)

modelsummary::modelsummary(list(m1, m2, m3), stars = TRUE)
```
