---
title: "The Effect of Historical Treatment on Population Density: France-Germany Border Region"
format: html
---

## Introduction

This notebook analyses the relationship between a historical treatment variable and present-day population density in the France-Germany border region. The dataset contains LAU-level regions straddling the French-German border, with a binary indicator (`treatment`) marking which regions received a Roman road, the historical treatment of interest.

We will:

1. Load and map the spatial data.
2. Filter to regions with a valid treatment assignment.
3. Estimate a bivariate OLS regression of population density on the treatment indicator.
4. Display the results in a publication-style table.

## Step 1: Load packages and data

```{r}
library(sf)
library(tidyverse)
```

We read in the France-Germany shapefile. Each row is a LAU region with spatial geometry, a treatment indicator (`treatment`), and population density (`POP_DENS_2016`).

```{r}
fr_gr <- sf::st_read("france_germany_updated.geojson")
```

A quick map to get a sense of the geography:

```{r}
fr_gr |>
  ggplot() +
  geom_sf()
```

## Step 2: Filter to regions with a treatment assignment

Some regions have `NA` for `treatment` — these are regions where treatment status is unknown or not applicable (e.g. regions far from the border). We drop them so that all three models are estimated on the same sample.

```{r}
data <- fr_gr |>
  filter(!is.na(treatment))

data |> 
  ggplot(aes(fill=treatment)) + geom_sf()
```

## Step 3: Estimate the regression model

We use `fixest::feols()` to estimate a bivariate OLS regression of population density on the treatment indicator.

```{r}
library(fixest)

m1 <- fixest::feols(POP_DENS_2016 ~ treatment, data = data)
```

## Step 4: Display results in a regression table

`modelsummary()` displays the model results. `stars = TRUE` adds the conventional significance stars (*, **, ***).

```{r}
library(modelsummary)

modelsummary::modelsummary(list(m1), stars = TRUE)
```
