---
title: "Synthetic Control: The Economic Impact of German Reunification"
format: html
---

## Introduction

The **synthetic control method** is a way to estimate the causal effect of a policy when only a small number of units are treated (e.g. a handful of states or countries). The idea is to build a weighted combination of untreated units — a *synthetic* version of each treated unit — that closely matches its pre-treatment trajectory. After the treatment, any divergence between the treated unit and its synthetic counterpart is attributed to the policy.

This notebook replicates the classic study of Abadie, Diamond & Hainmueller (2015), which estimates the economic impact of German Reunification (1990) on West Germany's GDP per capita using the `scpi` package.

We will:

1. Load and prepare the data.
2. Set up the synthetic control problem with `scdata()`.
3. Estimate the treatment effect with `scest()`.
4. Add uncertainty estimates (confidence intervals) with `scpi()`.

## Step 1: Load packages and data

```{r}
library(tidyverse)
library(scpi)
```

The `scpi` package ships with the `scpi_germany` dataset containing annual economic indicators for 17 OECD countries from 1960 to 2003. The treated unit is **West Germany**, which was reunified with East Germany in 1990.

```{r}
data("scpi_germany")
```

## Step 2: Set up the synthetic control problem

`scdata()` prepares the data structure that `scpi` needs. The key arguments are:

- `id.var`, `time.var` — which columns identify units and time periods.
- `outcome.var` — the variable we want to predict/compare (`gdp`).
- `period.pre` / `period.post` — pre- and post-treatment time periods.
- `unit.tr` — the treated unit (`"West Germany"`).
- `unit.co` — the donor pool of untreated countries.
- `features` — variables used to match the treated unit to its synthetic in the pre-period.
- `cov.adj` — additional adjustments applied to the matching (a constant and a linear trend).

```{r}
donors <- unique(scpi_germany$country)
donors <- donors[donors != "West Germany"]

df <- scdata(
  scpi_germany,
  id.var      = "country",
  outcome.var = "gdp",
  time.var    = "year",
  period.pre  = 1960:1989,
  period.post = 1990:2003,
  unit.tr     = "West Germany",
  unit.co     = donors,
  features    = c("gdp", "infrate", "trade", "schooling", "industry"),
  cov.adj     = list(c("constant", "trend"))
)
```

## Step 3: Estimate the treatment effect

`scest()` finds the optimal weights for the donor countries. The constraint `w.constr = list("name" = "simplex")` forces the weights to be non-negative and sum to one — so the synthetic control is a proper convex combination of donors, not an extrapolation.

`scplot()` then shows the actual outcome alongside the synthetic counterfactual.

```{r}
res <- scest(df, w.constr = list("name" = "simplex"))
scplot(res)
```

## Step 4: Add uncertainty with bootstrapped confidence intervals

Point estimates alone do not tell us whether the gap between treated and synthetic is statistically meaningful. `scpi()` adds inference via a Gaussian resampling procedure (`e.method = "gaussian"`), run in parallel across 4 cores with 50 simulations.

`scplot(..., type = "series")` shows the time series with confidence bands.

```{r}
respi <- scpi(df,
              w.constr = list("name" = "simplex"),
              cores    = 4,
              sims     = 50,
              e.method = "gaussian")

scplot(respi)
```

> **Note:** Increase `sims` (e.g. to 200 or 1000) for more stable confidence intervals in a real analysis. 50 is used here for speed.
