---
title: "Synthetic Control: The Effect of Smoking Laws on Cigarette Sales"
format: html
---

## Introduction

The **synthetic control method** is a way to estimate the causal effect of a policy when only a small number of units are treated (e.g. a handful of states or countries). The idea is to build a weighted combination of untreated units — a *synthetic* version of each treated unit — that closely matches its pre-treatment trajectory. After the treatment, any divergence between the treated unit and its synthetic counterpart is attributed to the policy.

This notebook uses a fictitious extension of the classic California smoking-law dataset. We assign anti-smoking legislation to **four US states** at different points in time and estimate its effect on cigarette sales using the `scpi` package, which supports **multiple treated units** with **staggered adoption**.

We will:

1. Load and prepare the data, defining when each state was treated.
2. Set up the synthetic control problem with `scdataMulti()`.
3. Estimate the treatment effect with `scest()`.
4. Add uncertainty estimates (confidence intervals) with `scpi()`.

## Step 1: Load packages and data

```{r}
library(tidyverse)
library(scpi)
```

The `scpi` package ships with a `smoking` dataset containing annual cigarette sales (`cigsale`) and retail prices (`retprice`) for US states. We load it with `data()`.

```{r}
data("smoking")
```

## Step 2: Define treatment timing

In reality, California is the only treated state in the original dataset. Here we fictitiously assign smoking-law adoption to four states, each at a different year:

| State      | Treated from |
|------------|-------------|
| California | 1976        |
| Alabama    | 1978        |
| Arkansas   | 1981        |
| Iowa       | 1977        |

We create a binary `treatment` column that equals 1 for a state in the years *after* it adopted the law, and 0 otherwise.

```{r}
smoking <- smoking |>
  mutate(treatment = if_else(
    (state == "Iowa"       & year > 1976) |
    (state == "Arkansas"   & year > 1980) |
    (state == "Alabama"    & year > 1977) |
    (state == "California" & year > 1975),
    1, 0
  ))
```

## Step 3: Set up the synthetic control problem

`scdataMulti()` prepares the data structure that `scpi` needs. The key arguments are:

- `id.var`, `time.var` — which columns identify units and time periods.
- `outcome.var` — the variable we want to predict/compare (`cigsale`).
- `treatment.var` — the 0/1 treatment indicator we just created.
- `donors.est` — the pool of untreated states used to build the synthetic controls.
- `features` — variables used to match treated units to their synthetics in the pre-period (here: sales and retail price).
- `cov.adj` — additional adjustments applied to the matching (a constant and a linear trend).
- `effect = "time"` — the most common choice; estimates a separate treatment effect for each post-treatment period.

```{r}
df <- scdataMulti(
  smoking,
  id.var       = "state",
  outcome.var  = "cigsale",
  treatment.var = "treatment",
  time.var     = "year",
  constant     = TRUE,
  donors.est   = list(c("North Dakota", "New Hampshire", "Oklahoma", "Tennessee")),
  features     = list(c("cigsale", "retprice")),
  cov.adj      = list(c("constant", "trend")),
  effect       = "time"
)
```

## Step 4: Estimate the treatment effect

`scest()` finds the optimal weights for the donor states. The constraint `w.constr = list("name" = "simplex")` forces the weights to be non-negative and sum to one — so the synthetic control is a proper convex combination of donors, not an extrapolation.

`scplotMulti()` then produces a plot for each treated state, showing the actual outcome alongside the synthetic counterfactual.

```{r}
res <- scest(df, w.constr = list("name" = "simplex"))
scplotMulti(res)
```

## Step 5: Add uncertainty with bootstrapped confidence intervals

Point estimates alone do not tell us whether the gap between treated and synthetic is statistically meaningful. `scpi()` adds inference via a Gaussian resampling procedure (`e.method = "gaussian"`), run in parallel across 4 cores with 50 simulations.

`scplotMulti(..., type = "series")` shows the time series with confidence bands for each treated state.

```{r}
respi <- scpi(df,
              w.constr = list("name" = "simplex"),
              cores    = 4,
              sims     = 50,
              e.method = "gaussian")

scplotMulti(respi, type = "series")
```

> **Note:** Increase `sims` (e.g. to 200 or 1000) for more stable confidence intervals in a real analysis. 50 is used here for speed.
