---
title: "Historical Persistence"
subtitle: "Applied Economics Research Course"
author: "Bas Machielsen"
institute: "Utrecht University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=F}
library(tidyverse); library(pacman); library(sf); library(terra); library(stars)
knitr::opts_chunk$set(warning=F, message=F, fig.retina = 3)
```


class: inverse, center, middle

# Some Common Data Sources

---

# Data Sources

- [Global Agro-Ecological Zones](https://gaez.fao.org/)
- [Gridded population of the world](https://cmr.earthdata.nasa.gov/search/concepts/C1597159135-SEDAC.html)
- [Gridded population of the Netherlands](https://www.pdok.nl/introductie/-/article/cbs-bevolkingsspreiding-population-distribution-)
- [Road Network Netherlands](https://www.pdok.nl/introductie/-/article/nationaal-wegen-bestand-nwb-wegen)
- [Worldwide Biodiversity Data](https://biodiversitymapping.org/index.php/download/)
- [Urban Land Cover Data (Europe only)](https://sdi.eea.europa.eu/catalogue/geoss/eng/catalog.search#/metadata/949683b7-5795-4c72-845f-77d049010649)
- [Urban Land Cover Data](https://livingatlas.arcgis.com/landcoverexplorer/#mapCenter=6.025%2C52.371%2C8&mode=step&timeExtent=2017%2C2022&year=2017)
- [Biodiversity datasets (Europe only)](https://sdi.eea.europa.eu/catalogue/biodiversity/eng/catalog.search#/search?isTemplate=n&resourceTemporalDateRange=%7B%22range%22:%7B%22resourceTemporalDateRange%22:%7B%22gte%22:null,%22lte%22:null,%22relation%22:%22intersects%22%7D%7D%7D&sortBy=relevance&sortOrder=&query_string=%7B%22th_httpinspireeceuropaeutheme-theme_tree.key%22:%7B%22http:%2F%2Finspire.ec.europa.eu%2Ftheme%2Fsd%22:true,%22http:%2F%2Finspire.ec.europa.eu%2Ftheme%2Fhb%22:true%7D,%22resourceType%22:%7B%22dataset%22:true%7D%7D&from=1&to=30)


- Some datasets might require registration at the proprietor

---

class: inverse, center, middle

# Some Common Data Wrangling Operations

---

# Import `.h5` nightlights data 

- Some nightlights data is structured as `.h5` data

- These are "layered" raster files that contain a potentially large amount of variables

- I will demonstrate how this works

- Furthermore, I will also show how to _combine_ different raster files

---

# Import `.h5` data

- A `.h5` dataset can be imported in the following way:
  - Load the `rhdf5` library:
  
```{r warning=FALSE, message=FALSE, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("rhdf5")
```

  - Extract the metadata, and read where the latitude, longitude data are contained

```{r}
library(rhdf5)
meta <- rhdf5::h5ls('VNP46A4.A2022001.h18v03.001.2023082112129.h5')
lat <- rhdf5::h5read('VNP46A4.A2022001.h18v03.001.2023082112129.h5',
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lat')
lon <- rhdf5::h5read('VNP46A4.A2022001.h18v03.001.2023082112129.h5',
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lon')
```

---

# Import `.h5` data

- Then, select one layer of the map you are interested in
  - I also import a shapefile of the Netherlands because I want to pay attention to this part
  - Set the `extent` of a raster to the latitude and longitude data
  - Set the `crs` of a raster to WGS84 (the default projection of the NASA VIIRS data)
  
```{r}
netherlands <- geodata::gadm("Netherlands", path="./") |> st_as_sf()
raster <- terra::rast('VNP46A4.A2022001.h18v03.001.2023082112129.h5') |>
  terra::subset("AllAngle_Composite_Snow_Covered")

ext(raster) <- c(min(lon), max(lon), min(lat), max(lat))
crs(raster) <- crs('wgs84')
```

---

# Import the `.h5` data

- Now, the data is usable in principle
- However, we do some post-processing:
  - We first mask (confine) the data set to the extent (overlap) with the Netherlands
  - Then we crop it so as to remove unnecessary `NA` data
  - Finally, we filter out the default sea level 

```{r}
masked_raster <- terra::mask(raster, netherlands)
netherlands_nightlights <- crop(masked_raster, netherlands)
netherlands_nightlights[netherlands_nightlights == 65535] = NA
```

---

# Plot the outcome

- Finally, we can plot the outcome

```{r, fig.align='center', fig.height=6}
terra::plot(sqrt(netherlands_nightlights))
```

---

# Combine various raster datasets

- By writing a function, you can also combine several datasets:

```{r eval=FALSE}
combine_and_clean_rasters <- function(directory_with_rasters,
                                      type="AllAngle_Composite_Snow_Covered"){
  # Find all the datasets in the directory
  nms <- list.files(directory_with_rasters, pattern='\\.h5')
  files <- paste0(directory_with_rasters, nms)
  # Pick one type of map
  rasters <- map(files, ~ terra::rast(.x) |>
                   terra::subset(type))
  # Extract lat, lon
  lat <- map(files, ~ rhdf5::h5read(.x,
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lat'))
  lon <- map(files, ~ rhdf5::h5read(.x,
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lon'))
  rasters <- imap(rasters, ~ { 
  ext(.x) <- c(min(lon[[.y]]), max(lon[[.y]]), min(lat[[.y]]), max(lat[[.y]])) 
  .x  })
  # Convert to correct CRS
  rasters <- map(rasters, ~ { crs(.x) <- crs('wgs84'); .x })
  out <- do.call(terra::mosaic, rasters)
  return(out)
}

```

---

# Road Density

- Now, I show how to combine the `roads` data to construct density

```{r, echo=T, results='hide'}
netherlands <- geodata::gadm("Netherlands", level=2, path="./") |> st_as_sf()
roads <- st_read('./nwb_hoofdwegen.gpkg') |> st_transform(crs=crs(netherlands)) |> select(geom)
```

```{r fig.width=4, fig.height=4, fig.align='center'}
roads |> ggplot() + geom_sf()
```

---

# Buffer the roads

- I build a buffer of 10 meters around each road
  - Rasterize it
  - Then set all values where there is a road to 1 (all roads are equal)
- Finally, plot it:

```{r, warning=FALSE, fig.height=3.5, fig.align='center'}
roads_buffer <- st_buffer(roads, dist=10)
roads_raster <- st_rasterize(roads_buffer,
                             options = "ALL_TOUCHED=TRUE")

roads_raster[roads_raster > 0] = 1; roads_raster[is.na(roads_raster)] = 0
final_raster <- terra::rast(roads_raster); terra::plot(final_raster)
```

---

# Aggregate the roads to a shapefile

- We already know how to do this - we can use the `extract` function from `terra`

```{r}
values <- terra::extract(final_raster, netherlands) |> 
  janitor::clean_names()

road_density <- values |> 
  group_by(id) |> 
  summarize(mean_roads = mean(id_2, na.rm=T))
```

- Finally, we merge the road density to the shapefile:

```{r}
nl_roads <- netherlands |> 
  mutate(ID=row_number()) |>
  left_join(road_density, by=c("ID" = "id"))
```

---

# Plot the output

- We end up with a road density per municipality:

```{r, fig.width=4, fig.height=4, fig.align='center'}
nl_roads |> 
  ggplot(aes(fill=mean_roads)) + 
  geom_sf() +
  scale_fill_viridis_c(option='F')
```


---

class: inverse, center, middle

# Land Cover Per Municipality

---

# Land Cover

- Download a map from e.g. [here](https://land.copernicus.eu/en/products)
  - I downloaded the CORINE Land Cover 2018 [Here](https://land.copernicus.eu/en/products/corine-land-cover)
- Import it (it is hidden in a .zip within a .zip)
- And put everything in the same projection:

```{r}
netherlands <- geodata::gadm("Netherlands", level=2, path="./") |> st_as_sf()
land_cover <- terra::rast('./u2018_clc2018_v2020_20u1_raster100m/DATA/U2018_CLC2018_V2020_20u1.tif')
netherlands <- st_transform(netherlands, st_crs(land_cover))
```

---

# Aggregate categories per polygon

- Implement the following `for`-loop:

```{r}
# Create an empty data frame to store results
results <- data.frame()

# Loop through each polygon
for (i in 1:nrow(netherlands)) {
  polygon <- netherlands[i,]
  # Clip the raster to the current polygon
  raster_clipped <- terra::crop(land_cover, polygon)
  # Calculate the area covered by each value within the polygon
  extracted_values <- extract(raster_clipped, polygon)
  out <- extracted_values |> 
    group_by(LABEL3) |> count() |> ungroup() |>
    mutate(tot = sum(n))   |>
    group_by(LABEL3) |>
    mutate(percentage = n/tot) |> dplyr::select(-c(n, tot)) |>
    pivot_wider(names_from=LABEL3, values_from = percentage)
  
  # Add polygon ID and coverage percentages to results
  results <- bind_rows(results, out)
}
```

---

# Output

- What you end up with is this:

```{r}
results |> 
  select(c(1:4)) |>
  head(5)
```

---

# Merge

- Now, we can put this together in Netherlands by merging on the basis of row number..:
  - Beforehand, we fill in `NA` observations with zero:
  
```{r}
results <- results |> mutate(across(everything(), ~ replace_na(.x, 0)))
netherlands <- netherlands |> bind_cols(results)
```

- Let's plot the most urban municipalities:

```{r fig.align='center', fig.width=5, fig.height=3}
netherlands |>
  ggplot(aes(fill=`Discontinuous urban fabric`)) + geom_sf()
```


