---
title: "Historical Persistence"
subtitle: "Applied Economics Research Course"
author: "Bas Machielsen"
institute: "Utrecht University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=F}
library(tidyverse); library(pacman); library(sf); library(terra)
knitr::opts_chunk$set(warning=F, message=F, fig.retina = 3)
```

class: inverse, center, middle

# Some Common Data Wrangling Operations

---

# Import `.h5` nightlights data 

- Some nightlights data is structured as `.h5` data

- These are "layered" raster files that contain a potentially large amount of variables

- I will demonstrate how this works

- Furthermore, I will also show how to _combine_ different raster files

---

# Import `.h5` data

- A `.h5` dataset can be imported in the following way:
  - Load the `rhdf5` library:
  - Extract the metadata, and read where the latitude, longitude data are contained

```{r}
library(rhdf5)
meta <- rhdf5::h5ls('VNP46A4.A2022001.h18v03.001.2023082112129.h5')
lat <- rhdf5::h5read('VNP46A4.A2022001.h18v03.001.2023082112129.h5',
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lat')
lon <- rhdf5::h5read('VNP46A4.A2022001.h18v03.001.2023082112129.h5',
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lon')
```

---

# Import `.h5` data

- Then, select one layer of the map you are interested in
  - I also import a shapefile of the Netherlands because I want to pay attention to this part
  - Set the `extent` of a raster to the latitude and longitude data
  - Set the `crs` of a raster to WGS84 (the default projection of the NASA VIIRS data)
  
```{r}
netherlands <- geodata::gadm("Netherlands", path="./") |> st_as_sf()
raster <- terra::rast('VNP46A4.A2022001.h18v03.001.2023082112129.h5') |>
  subset("AllAngle_Composite_Snow_Covered")

ext(raster) <- c(min(lon), max(lon), min(lat), max(lat))
crs(raster) <- crs('wgs84')
```

---

# Import the `.h5` data

- Now, the data is usable in principle
- However, we do some post-processing:
  - We first mask (confine) the data set to the extent (overlap) with the Netherlands
  - Then we crop it so as to remove unnecessary `NA` data
  - Finally, we filter out the default sea level 

```{r}
masked_raster <- terra::mask(raster, netherlands)
netherlands_nightlights <- crop(masked_raster, netherlands)
netherlands_nightlights[netherlands_nightlights == 65535] = NA
```

---

# Plot the outcome

- Finally, we can plot the outcome

```{r}
plot(sqrt(netherlands_nightlights))
```

---

# Combine various raster datasets

- By writing a function, you can also combine several datasets:

```{r}
combine_and_clean_rasters <- function(directory_with_rasters,
                                      type="AllAngle_Composite_Snow_Covered"){
  
  # Find all the datasets in the directory
  nms <- list.files(directory_with_rasters, pattern='\\.h5')
  files <- paste0(directory_with_rasters, nms)
  
  # Pick one type of map
  rasters <- map(files, ~ terra::rast(.x) |>
                   subset(type))
  # Extract lat, lon
  lat <- map(files, ~ rhdf5::h5read(.x,
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lat'))
  lon <- map(files, ~ rhdf5::h5read(.x,
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lon'))
  
  # Add lat, lon
  rasters <- imap(rasters, ~ { 
  ext(.x) <- c(min(lon[[.y]]), max(lon[[.y]]), min(lat[[.y]]), max(lat[[.y]])) 
  .x
  })
  
  # Convert to correct CRS
  rasters <- map(rasters, ~ { crs(.x) <- crs('wgs84'); .x })
  
  out <- do.call(terra::mosaic, rasters)
  return(out)
  
}

```


--- 

# Convert shapefiles to raster

---


