---
title: "Historical Persistence"
subtitle: "Applied Economics Research Course"
author: "Bas Machielsen"
institute: "Utrecht University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=F}
library(tidyverse); library(pacman); library(sf); library(terra); library(stars)
knitr::opts_chunk$set(warning=F, message=F, fig.retina = 3)
```

class: inverse, center, middle

# Some Common Data Wrangling Operations

---

# Import `.h5` nightlights data 

- Some nightlights data is structured as `.h5` data

- These are "layered" raster files that contain a potentially large amount of variables

- I will demonstrate how this works

- Furthermore, I will also show how to _combine_ different raster files

---

# Import `.h5` data

- A `.h5` dataset can be imported in the following way:
  - Load the `rhdf5` library:
  
```{r warning=FALSE, message=FALSE, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("rhdf5")
```

  - Extract the metadata, and read where the latitude, longitude data are contained

```{r}
library(rhdf5)
meta <- rhdf5::h5ls('VNP46A4.A2022001.h18v03.001.2023082112129.h5')
lat <- rhdf5::h5read('VNP46A4.A2022001.h18v03.001.2023082112129.h5',
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lat')
lon <- rhdf5::h5read('VNP46A4.A2022001.h18v03.001.2023082112129.h5',
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lon')
```

---

# Import `.h5` data

- Then, select one layer of the map you are interested in
  - I also import a shapefile of the Netherlands because I want to pay attention to this part
  - Set the `extent` of a raster to the latitude and longitude data
  - Set the `crs` of a raster to WGS84 (the default projection of the NASA VIIRS data)
  
```{r}
netherlands <- geodata::gadm("Netherlands", path="./") |> st_as_sf()
raster <- terra::rast('VNP46A4.A2022001.h18v03.001.2023082112129.h5') |>
  terra::subset("AllAngle_Composite_Snow_Covered")

ext(raster) <- c(min(lon), max(lon), min(lat), max(lat))
crs(raster) <- crs('wgs84')
```

---

# Import the `.h5` data

- Now, the data is usable in principle
- However, we do some post-processing:
  - We first mask (confine) the data set to the extent (overlap) with the Netherlands
  - Then we crop it so as to remove unnecessary `NA` data
  - Finally, we filter out the default sea level 

```{r}
masked_raster <- terra::mask(raster, netherlands)
netherlands_nightlights <- crop(masked_raster, netherlands)
netherlands_nightlights[netherlands_nightlights == 65535] = NA
```

---

# Plot the outcome

- Finally, we can plot the outcome

```{r, fig.align='center', fig.height=6}
terra::plot(sqrt(netherlands_nightlights))
```

---

# Combine various raster datasets

- By writing a function, you can also combine several datasets:

```{r eval=FALSE}
combine_and_clean_rasters <- function(directory_with_rasters,
                                      type="AllAngle_Composite_Snow_Covered"){
  # Find all the datasets in the directory
  nms <- list.files(directory_with_rasters, pattern='\\.h5')
  files <- paste0(directory_with_rasters, nms)
  # Pick one type of map
  rasters <- map(files, ~ terra::rast(.x) |>
                   terra::subset(type))
  # Extract lat, lon
  lat <- map(files, ~ rhdf5::h5read(.x,
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lat'))
  lon <- map(files, ~ rhdf5::h5read(.x,
                     name='/HDFEOS/GRIDS/VIIRS_Grid_DNB_2d/Data Fields/lon'))
  rasters <- imap(rasters, ~ { 
  ext(.x) <- c(min(lon[[.y]]), max(lon[[.y]]), min(lat[[.y]]), max(lat[[.y]])) 
  .x  })
  # Convert to correct CRS
  rasters <- map(rasters, ~ { crs(.x) <- crs('wgs84'); .x })
  out <- do.call(terra::mosaic, rasters)
  return(out)
}

```

---

# Road Density

- Now, I show how to combine the `roads` data to construct density

```{r, echo=T, results='hide'}
netherlands <- geodata::gadm("Netherlands", level=2, path="./") |> st_as_sf()
roads <- st_read('./nwb_hoofdwegen.gpkg') |> st_transform(crs=crs(netherlands)) |> select(geom)
```

```{r fig.width=4, fig.height=4, fig.align='center'}
roads |> ggplot() + geom_sf()
```

---

# Buffer the roads

- I build a buffer of 10 meters around each road
  - Rasterize it
  - Then set all values where there is a road to 1 (all roads are equal)
- Finally, plot it:

```{r, warning=FALSE, fig.height=3.5, fig.align='center', cache=TRUE}
roads_buffer <- st_buffer(roads, dist=10)
roads_raster <- st_rasterize(roads_buffer,
                             options = "ALL_TOUCHED=TRUE")

roads_raster[roads_raster > 0] = 1; roads_raster[is.na(roads_raster)] = 0
final_raster <- rast(roads_raster); terra::plot(final_raster)
```

---

# Aggregate the roads to a shapefile

- We already know how to do this - we can use the `extract` function from `terra`

```{r}
values <- terra::extract(final_raster, netherlands)
road_density <- values |> 
  group_by(ID) |> 
  summarize(mean_roads = mean(lyr.1, na.rm=T))
```

- Finally, we merge the road density to the shapefile:

```{r}
nl_roads <- netherlands |> 
  mutate(ID=row_number()) |>
  left_join(road_density, by="ID")
```

---

# Plot the output

- We end up with a road density per municipality:

```{r, fig.width=4, fig.height=4, fig.align='center'}
nl_roads |> 
  ggplot(aes(fill=mean_roads)) + 
  geom_sf() +
  scale_fill_viridis_c(option='F')
```

