---
title: "Historical Persistence & Geospatial Economics"
subtitle: "Applied Economics Research Course"
author: "Bas Machielsen"
aspectratio: 169
format: beamer
header-includes:
  - \usepackage{underscore}
execute:
  freeze: auto
---

```{r setup}
#| include: false
library(tidyverse); library(sf); library(stars); library(here); library(terra)
sf_use_s2(FALSE)
# terra masks dplyr::select and dplyr::filter — restore dplyr versions
select <- dplyr::select
filter <- dplyr::filter
```

# Introduction

## Why Spatial Data Wrangling?

- Your research project starts with `.geojson` / `.shp` files
- Before any regression, you need to be able to:
  - **Load** spatial data and understand its structure
  - **Clean and filter** observations
  - **Join** datasets that live in different files
  - **Aggregate** fine-grained data to your unit of analysis
- This lecture walks through all four steps using the **course example datasets**

## Example Dataset

```{r load_fg}
#| cache: true
#| output-location: slide
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
fg <- st_read(here("data", "france_germany", "france_germany_updated.geojson"),
              quiet = TRUE)
fg |> ggplot(aes(fill = POP_DENS_2016)) +
  geom_sf(linewidth = 0.05) +
  scale_fill_viridis_c(trans = "log", name = "Pop. Density (log)") +
  labs(title = "France--Germany Border Region")
```

## Two Data Models

| Feature | **Vector** | **Raster** |
|---|---|---|
| Representation | Points / Lines / Polygons | Regular grid of cells |
| Format | `sf` data.frame | Image-like matrix |
| Example | Municipality boundaries | Satellite imagery, climate |
| R packages | `sf` | `stars`, `terra`, `raster` |

- **Vector** suits analysis where your unit is a well-defined geographic entity (municipality, country)
- **Raster** suits analysis where your unit is a cell at fixed resolution (0.05° × 0.05°)
- In practice: you often *convert* between the two

## Coordinate Reference Systems

- Every spatial object has a **CRS** — a mathematical system that maps coordinates to locations on Earth
- CRS matters because:
  - Distance calculations assume a specific unit (meters vs. degrees)
  - Two datasets in different CRS produce nonsense when combined
- Use `st_crs()` to inspect, `st_transform()` to reproject

```{r crs_demo}
#| echo: true
st_crs(fg)$input   # inspect
fg_wgs84 <- st_transform(fg, crs = 4326)   # reproject to WGS84
st_crs(fg_wgs84)$input
```

- Rule of thumb: always check CRS before joining two spatial datasets

# Part 1: Computing on Vector Data

## Loading Data

- Use `st_read()` + `here()` to load course GeoJSON files
- `here()` always resolves from the project root — no hard-coded paths

```{r load_nl}
#| echo: true
fg  <- st_read(here("data", "france_germany", "france_germany_updated.geojson"),
               quiet = TRUE)
nl  <- st_read(here("data", "netherlands", "netherlands_roman_updated.geojson"),
               quiet = TRUE)
```

- `class()` confirms the object is both `sf` **and** a `data.frame`:

```{r class_fg}
#| echo: true
class(fg)
```

## Inspecting sf Objects {shrink=20}

- `sf` objects behave exactly like a `data.frame` with one extra geometry column

```{r inspect}
head(fg, 3)
```

## Inspecting sf Objects (cont.)

- Geometry-specific helpers give spatial metadata:

```{r inspect2}
#| echo: true
#| size: "\\scriptsize"
st_geometry_type(fg) |> unique()
st_bbox(fg)
glimpse(fg)
```

## dplyr Verbs Work as Normal

- `filter()`, `mutate()`, `select()` all work unchanged on `sf` objects

```{r dplyr_demo}
#| echo: true
# Keep only French municipalities, add log population density
france <- fg |>
  filter(treatment == 0 | treatment == 1) |>     # all rows — just to show filter
  mutate(log_pop = log(POP_DENS_2016 + 1)) |>
  select(log_pop, treatment, geometry)

france |> head(3)
```

## dplyr Verbs: Plot Result

```{r dplyr_plot}
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
france |>
  ggplot(aes(fill = log_pop)) +
  geom_sf(linewidth = 0.05) +
  scale_fill_viridis_c(name = "log(Pop. Density)") +
  labs(title = "Log Population Density -- France/Germany Border")
```

## CRS Operations

- Always check CRS before joining datasets:

```{r crs_check}
#| echo: true
st_crs(fg)$input
st_crs(nl)$input
```

- Reproject with `st_transform()`:

```{r crs_transform}
# Project to ETRS89 / LAEA (metric CRS, EPSG:3035) for distance calculations
fg_metric <- st_transform(fg, crs = 3035)
nl_metric <- st_transform(nl, crs = 3035)
```

- **Warning**: mixing CRS silently produces wrong geometries and distances

## Geometric Operations

- `sf` supports set operations on geometries:
  - `st_buffer(x, dist)` — expand each geometry outward by `dist` metres
  - `st_union(x)` — merge all geometries into one
  - `st_intersection(x, y)` — keep the overlapping parts
  - `st_difference(x, y)` — remove the part of x that overlaps y

```{r geom_ops}
#| echo: true
# Small illustrative subset: two neighbouring municipalities
sub <- fg_metric |> slice(1:2)
buf <- st_buffer(sub, dist = 5000)     # 5 km buffer
uni <- st_union(sub)                   # merge the two polygons
```

## Geometric Operations Plot

```{r geom_plot}
#| fig-width: 5
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
ggplot() +
  geom_sf(data = buf, fill = "steelblue", alpha = 0.3) +
  geom_sf(data = sub, fill = "orange", alpha = 0.6) +
  labs(title = "Buffer (blue) around two municipalities (orange)")
```

## Computing Distances

- `st_centroid()` converts polygons to their centre points
- `st_distance()` returns a matrix of pairwise distances

```{r distances}
#| echo: true
# Centroids of first 5 municipalities
cents <- fg_metric |> slice(1:5) |> st_centroid()
# Distance matrix (meters, in metric CRS)
st_distance(cents) |> round(0)
```

## Distance to Roman Roads — Course Context

- The `france_germany_updated.geojson` already contains a distance-to-road variable
- Here is how one would build it from scratch:

```{r dist_concept}
#| eval: false
#| echo: true
# Pseudocode (road layer not included here)
centroids <- fg_metric |> st_centroid()
dist_matrix <- st_distance(centroids, roman_roads_metric)
fg_metric <- fg_metric |>
  mutate(min_dist = apply(dist_matrix, 1, min))
```

- The `treatment` variable in the course data is a binary indicator:
  `treatment == 1` if a Roman road passes through the municipality

## Spatial Joins: Concept

- `st_join(x, y, join = <predicate>)` enriches `x` with attributes from `y`
  - based on a **geometric relationship** between their geometries, not a shared key

| Predicate | Meaning |
|---|---|
| `st_intersects` | x and y share any point (default) |
| `st_within` | x lies entirely inside y |
| `st_covered_by` | x is covered by y (boundary included) |
| `st_nearest_feature` | match x to closest y |

- Result: one row per match (may duplicate rows if many-to-many)
- Use `largest = TRUE` to keep only the y with largest overlap per x row

## Spatial Joins: Example

- Add a NUTS-2 region label to each municipality in our course data

```{r nuts_download}
#| echo: true
# Download NUTS-2 regions for France (small file, ~200 KB)
library(giscoR)
nuts2 <- giscoR::gisco_get_nuts(
  year = "2021", nuts_level = "2",
  country = "FR", resolution = "20"
)
nuts2 <- nuts2 |> select(NUTS_ID, NAME_LATN)
```

## Spatial Joins: Example

```{r nuts_join}
#| echo: true
fg_nuts <- st_join(
  st_transform(fg, st_crs(nuts2)),
  nuts2,
  join = st_within
)
fg_nuts |> select(NUTS_ID, NAME_LATN, POP_DENS_2016) |> head(3)
```

## Spatial Joins: Plot

```{r nuts_plot}
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
fg_nuts |>
  ggplot(aes(fill = NAME_LATN)) +
  geom_sf(linewidth = 0.05, show.legend = FALSE) +
  labs(title = "Municipalities coloured by NUTS-2 region")
```

## Merging with Non-Spatial Data

- An `sf` object **is** a `data.frame` — `left_join()` works exactly as expected

```{r tabular_join}
#| echo: true
# Simulate a CSV with per-municipality GDP data
fake_gdp <- fg |>
  st_drop_geometry() |>
  slice(1:20) |>
  select(LAU_ID) |>
  mutate(gdp_index = runif(20, 80, 130))

fg_gdp <- left_join(fg |> slice(1:20), fake_gdp, by = "LAU_ID")
fg_gdp |> select(LAU_ID, POP_DENS_2016, gdp_index) |> head(3)
```

- The geometry column travels with the data — no special handling needed

## Putting It Together

```{r final_map}
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
fg |>
  mutate(treated = factor(treatment, labels = c("No Roman Road", "Roman Road"))) |>
  ggplot(aes(fill = treated)) +
  geom_sf(linewidth = 0.05, alpha = 0.8) +
  scale_fill_viridis_d(name = "Treatment") +
  labs(title = "Roman Road Treatment -- France/Germany Border Municipalities")
```

# Part 2: Computing on Raster Data

## What Is Raster Data?

- A raster is a **regular grid** of cells, each containing a value
- **Resolution** = the geographic size of one cell (e.g. 100 m × 100 m)
- Each cell has implicit coordinates determined by the grid's origin + resolution + CRS

```{r raster_schematic}
#| eval: false
# Conceptual structure of a raster:
# [row 1, col 1]  [row 1, col 2]  [row 1, col 3] ...
# [row 2, col 1]  ...
# Each cell stores one numeric value (or NA)
```

Common raster sources:

- **Satellite imagery**: RGB bands + derived indices (NDVI, NDWI)
- **Climate**: temperature, precipitation (WorldClim, ERA5)
- **Air quality**: NOx, PM2.5 concentration grids

## Loading Raster Data

- `stars::read_stars()` is the modern approach; `raster::raster()` is the classic

```{r load_raster}
#| echo: true
nox <- read_stars(here("data", "schools", "nox_avg_22.tif"))
nox
```

## Plot Raster Data

```{r plot_raster_base}
#| fig-width: 5
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
plot(nox, main = "NOx Concentration 2022 (Europe)")
```

## Raster Algebra

- Arithmetic works element-wise on the entire grid
- Supported: `+`, `-`, `*`, `/`, `sqrt()`, `log()`, `abs()`, logical operators

```{r raster_algebra}
#| echo: true
# Scale NOx values (divide by 1000 to convert units) and take log
nox_log <- log(nox + 1)

# Threshold: flag cells with high NOx
nox_high <- nox > quantile(nox[[1]], 0.9, na.rm = TRUE)
```

## Raster Algebra Result

```{r algebra_plot}
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
ggplot() +
  geom_stars(data = nox_log) +
  scale_fill_viridis_c(name = "log(NOx + 1)", na.value = "white") +
  labs(title = "Log NOx Concentration")
```

## High-Level Functions

- `st_warp()` — reproject a `stars` raster to a new CRS or resolution
- `st_crop()` — clip to a bounding box or polygon
- `aggregate()` — reduce resolution (e.g. average 10 × 10 cells → 1 cell)

```{r highlevel}
#| echo: true
# Download France outline to tempdir
france_outline <- geodata::gadm("France", level = 0, path = tempdir()) |>
  st_as_sf() |>
  st_transform(st_crs(nox))

# Crop and downsample
nox_crop   <- st_crop(nox, france_outline)
nox_coarse <- st_warp(nox_crop, cellsize = c(0.1, 0.1), crs=4326)
```

## Visualizing Raster with ggplot

```{r raster_ggplot}
#| fig-width: 7
#| fig-height: 3.5
#| fig-align: 'center'
#| output-location: slide
ggplot() +
  geom_stars(data = nox_coarse) +
  geom_sf(data = st_transform(france_outline, 4326), fill = NA, colour = "black", linewidth = 0.3) +
  scale_fill_viridis_c(name = "NOx (µg/m³)", na.value = "white") +
  labs(title = "Average NOx Concentration 2022 -- France (downsampled)")
```

## Using `stars` Idioms: Crop to Polygon

- `st_crop()` with a polygon clips the raster to its spatial extent
- Combine with `st_union()` to dissolve internal boundaries first

```{r stars_crop}
#| echo: true
nox_france <- nox |>
  st_transform(st_crs(france_outline)) |>
  st_crop(st_union(france_outline))

nox_france
```

## Crop Result

```{r stars_crop_plot}
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
ggplot() +
  geom_stars(data = nox_france) +
  scale_fill_viridis_c(na.value = "white", name = "NOx") +
  labs(title = "NOx cropped to France boundary")
```

# Part 3: Aggregate Raster to Vector

## The Task

- **Goal**: Assign one climate value per municipality polygon
- **Why**: Municipalities are your unit of analysis; temperature or precipitation is a control variable stored as a raster
- **Workflow**:
  1. Load municipality polygons (vector)
  2. Download climate raster
  3. Align CRS between vector and raster
  4. Extract raster values per polygon and summarise (e.g. mean)

This is how you would add a "mean annual temperature" control to the France/Germany dataset.

## Download Climate Raster

```{r climate_download}
#| echo: true
# WorldClim monthly average temperature for France — downloads to tempdir()
tavg_rast <- geodata::worldclim_country(
  "France", var = "tavg", path = tempdir()
)
# tavg_rast is a SpatRaster with 12 layers (one per month)
tavg_rast
```

## Climate Plot

```{r climate_plot}
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
# Plot July (layer 7) using terra
terra::plot(tavg_rast[[7]], main = "Average Temperature July -- France")
```

## Align CRS

- Before extraction, **both objects must share a CRS**
- Common mistake: skip this step → values assigned to wrong polygons

```{r align_crs}
#| echo: true
# Check CRS
crs(tavg_rast)           # terra SpatRaster
st_crs(fg)$input         # sf data.frame

# Transform vector to match raster CRS
fg_proj <- st_transform(fg, crs = crs(tavg_rast))

# Verify they match
identical(st_crs(fg_proj)$proj4string,
          as.character(crs(tavg_rast, proj = TRUE)))
```

## Using `terra::extract()`

```{r terra_extract}
#| echo: true
# Extract mean July temperature per municipality polygon
values_july <- terra::extract(
  tavg_rast[[7]],        # raster layer
  vect(fg_proj),         # sf → SpatVector for terra
  fun = mean,
  na.rm = TRUE
)

fg_temp <- fg_proj |>
  mutate(mean_temp_july = values_july[, 2])

fg_temp |> select(mean_temp_july) |> head(4)
```

## `terra` Plot

```{r extract_plot}
#| fig-width: 7
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
fg_temp |>
  ggplot(aes(fill = mean_temp_july)) +
  geom_sf(linewidth = 0.05) +
  scale_fill_viridis_c(name = "°C") +
  labs(title = "Mean July Temperature per Municipality -- France/Germany")
```

## Using `stars::aggregate()`

- Alternative: convert raster to `stars`, then aggregate with `sf` polygons

```{r stars_aggregate}
#| echo: true
tavg_stars <- st_as_stars(tavg_rast[[7]])
tavg_stars <- st_transform(tavg_stars, st_crs(fg))

agg_result <- stats::aggregate(tavg_stars, fg, FUN = mean, na.rm = TRUE) |>
  st_as_sf() |>
  rename(mean_temp = 1)   # rename auto-generated layer column

agg_result |> head(3)
```

## `stars` Plot

```{r stars_agg_plot}
#| fig-width: 7
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
agg_result |>
  ggplot(aes(fill = mean_temp)) +
  geom_sf(linewidth = 0.05) +
  scale_fill_viridis_c(name = "°C") +
  labs(title = "Mean July Temperature -- stars::aggregate() approach")
```

# Part 4: Aggregate Vector to Raster

## The Task

- **Goal**: Replace irregular municipality polygons with a **regular spatial grid**
- **Why**: Some research designs use grid cells as the unit of analysis
  - Easier to merge datasets from different sources
  - Avoids Modifiable Areal Unit Problem (MAUP) from administrative boundaries
- **Workflow**:
  1. Create a regular grid over the study area
  2. Assign polygon attributes to overlapping grid cells (spatial join / aggregate)
  3. Plot and verify

## Creating a Grid

- `st_make_grid()` creates a grid of square (or hexagonal) cells

```{r make_grid}
#| cache: true
# 0.05° × 0.05° grid over the france_germany area
grid <- st_make_grid(fg, cellsize = c(0.05, 0.05), square = TRUE) |>
  st_as_sf() |>
  mutate(cell_id = row_number())

grid |> head(3)
```

## Grid Plot

```{r grid_plot}
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
#| output-location: slide
ggplot() +
  geom_sf(data = fg, fill = "steelblue", alpha = 0.4, linewidth = 0.05) +
  geom_sf(data = grid, fill = NA, colour = "black", linewidth = 0.2) +
  labs(title = "0.05° × 0.05° grid over France/Germany municipalities")
```

## Aggregating to Grid Cells

- Spatial join: for each grid cell, find which municipalities overlap, then average

```{r grid_aggregate}
#| echo: true
#| cache: true
# Spatial join: attach municipality attributes to intersecting grid cells
fg_grid_join <- st_join(grid, fg |> select(POP_DENS_2016, treatment))

# Average POP_DENS_2016 per grid cell (some cells overlap multiple municipalities)
fg_gridded <- fg_grid_join |>
  group_by(cell_id) |>
  summarise(
    POP_DENS_2016 = mean(POP_DENS_2016, na.rm = TRUE),
    treatment     = mean(treatment,     na.rm = TRUE),
    .groups       = "drop"
  )

```

## Plot the Gridded Result

```{r grid_final_plot}
#| cache: true
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
fg_gridded |>
  filter(!is.na(POP_DENS_2016)) |>
  ggplot(aes(fill = log(POP_DENS_2016 + 1))) +
  geom_sf(linewidth = 0.05) +
  scale_fill_viridis_c(name = "log(Pop. Density)") +
  labs(title = "Population Density Gridded to 0.05° × 0.05° Cells",
       subtitle = "Trade-off: lose polygon precision, gain spatial regularity")
```

- **Trade-off**: grid cells ignore administrative boundaries → some cells mix municipalities
- **Gain**: unit of analysis is consistent across datasets from different sources
